---
title: "Adding_Datasets"
output: html_document
---

Load libraries:
```{r}
library("glue") # Interpreted String Literals
library("here") # A Simpler Way to Find Your Files
library('tidyverse') # Easily Install and Load the 'Tidyverse'
library('usethis') # Automate Package and Project Setup
library('devtools') # Tools to Make Developing R Packages Easier
library('readr') # Read tables
library('readxl') # Read excel table 
```

```{r}
census_geographical_keys <- read_csv(paste0(here::here(), "/data-ignore-inputs/census_geographical_keys.csv"), 
    # col_types = cols(CTUID = col_factor(), 
    #     CTNAME = col_factor(), PR = col_factor(), 
    #     PCSD = col_factor(), PCSDNAME = col_factor(), 
    #     PCD = col_factor(), PCDNAME = col_factor(), 
    #     CMA = col_factor(), CMANAME = col_factor(), 
    #     PRNAME = col_factor())
    # 
    col_types = cols(.default = col_factor())) %>% 
  rename(PCSD = "CSDUID", 
         PCSDNAME = "CSDNAME", 
         PCD = "CDUID",
         PCDNAME = "CDNAME", 
         CMA = "CMAUID", 
         PR = "PRUID")
```


```{r}
cma_keys <- census_geographical_keys %>% 
  dplyr::distinct(CMA, .keep_all = TRUE) %>% 
  dplyr::select(CMA:PRNAME) %>% 
  filter(!is.na(CMA))

pcd_keys <- census_geographical_keys %>% 
  dplyr::distinct(PCD, .keep_all = TRUE) %>% 
  dplyr::select(PCD:PRNAME) 

ct_keys <- census_geographical_keys %>% 
  dplyr::distinct(CTUID, .keep_all = TRUE) %>% 
  dplyr::select(CTUID:PRNAME) %>% 
  filter(!is.na(CTUID))
```

# Census Testing Datasets

## Canada level

```{r}
# reading file
census_test_data <- read_csv(glue::glue(here::here(), 
                                        "/data-ignore-inputs/census_test_v5.csv"))
```

```{r}
# adding file to the package
usethis::use_data(census_test_data,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Toronto level

```{r}
# reading file
census_test_toronto <- read_csv(glue::glue(here::here(), 
                                  "/data-ignore-inputs/census_test_toronto_v2.csv"))
```

```{r}
# adding file to the package
usethis::use_data(census_test_toronto,
                  compress = "bzip2",
                  overwrite = TRUE)
```

# Impedance Functions

## Census Division Level 

```{r}
# reading file
pcd_impedance_functions <-read_csv(glue::glue(here::here(),                                               "/data-ignore-inputs/pcd_impedance_functions.csv"))

pcd_impedance_functions <- pcd_impedance_functions %>% 
  dplyr::select("PCD", "PwMode", "distribution", "est_1", "est_2") %>%
  rename("PwMode_label" = PwMode, Distribution = distribution) %>%
  mutate(PwMode_label = case_when(PwMode_label == "Car/motor" ~ "Car-motorized", 
                                  PwMode_label == "Transit" ~ "Public transit", 
                                  TRUE ~ PwMode_label),
         PwMode_label = factor(PwMode_label),
         PCD = factor(PCD), 
         est_2 = round(est_2, 5), 
         est_2 = case_when(Distribution == "exp" ~ NA, 
                          TRUE ~ est_2), 
         est_1 = round(est_1, 5),
         Distribution = case_when(Distribution == "exp" ~ "Exponential",
                                  Distribution == "gamma" ~ "Gamma",
                                  Distribution == "lnorm" ~ "Log-normal",
                                  Distribution == "norm" ~ "Normal",
                                  Distribution == "unif" ~ "Uniform"), 
         Distribution = factor(Distribution))

pcd_impedance_functions <- pcd_impedance_functions %>% 
  left_join(pcd_keys, by = "PCD")  %>% 
  dplyr::select(PCD, PCDNAME, "CMA", "CMANAME","PR", "PRNAME", "PwMode_label", "Distribution", "est_1", "est_2")
```


```{r}
# adding file to the package
usethis::use_data(pcd_impedance_functions,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Census Metropolitan Area Level 

```{r}
# reading file
cma_impedance_functions <- read_csv(glue::glue(here::here(), 
                      "/data-ignore-inputs/cma_impedance_functions.csv"))

cma_impedance_functions <- cma_impedance_functions %>% 
  dplyr::select("CMA", "PwMode", "distribution", "est_1", "est_2") %>%
  rename("PwMode_label" = PwMode, Distribution = distribution) %>%
  filter(CMA > 0, CMA < 996) %>% 
  mutate(PwMode_label = case_when(PwMode_label == "Car/motor" ~ "Car-motorized", 
                                  PwMode_label == "Transit" ~ "Public transit", 
                                  TRUE ~ PwMode_label),
         PwMode_label = factor(PwMode_label),
         CMA = factor(sprintf("%03d", as.numeric(CMA))), 
         est_2 = round(est_2, 5), 
         est_2 = case_when(Distribution == "exp" ~ NA, 
                          TRUE ~ est_2), 
         est_1 = round(est_1, 5),
         Distribution = case_when(Distribution == "exp" ~ "Exponential",
                                  Distribution == "gamma" ~ "Gamma",
                                  Distribution == "lnorm" ~ "Log-normal",
                                  Distribution == "norm" ~ "Normal",
                                  Distribution == "unif" ~ "Uniform"), 
         Distribution = factor(Distribution))

cma_impedance_functions <- cma_impedance_functions %>% 
  left_join(cma_keys, by = "CMA") %>% 
  dplyr::select("CMA", "CMANAME","PR", "PRNAME", "PwMode_label", "Distribution", "est_1", "est_2")
```


```{r}
# adding file to the package
usethis::use_data(cma_impedance_functions,
                  compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## Province Level 

```{r}
# reading file
pr_impedance_functions <- read_csv(glue::glue(here::here(), 
                                              "/data-ignore-inputs/pr_impedance_functions.csv"))

pr_impedance_functions <- pr_impedance_functions %>% 
  dplyr::select("Pr", 
                "CMA_type", 
                "PwMode", "distribution", "est_1", "est_2") %>%
  rename("PwMode_label" = PwMode, PRNAME = Pr, CMACATEG = CMA_type, 
         Distribution = distribution) %>%
  mutate(PwMode_label = case_when(PwMode_label == "Car/motor" ~ "Car-motorized", 
                                  PwMode_label == "Transit" ~ "Public transit", 
                                  TRUE ~ PwMode_label),
         PwMode_label = factor(PwMode_label),
         CMACATEG = factor(CMACATEG), 
         est_2 = round(est_2, 5), 
         est_2 = case_when(Distribution == "exp" ~ NA, 
                          TRUE ~ est_2), 
         est_1 = round(est_1, 5),
         Distribution = case_when(Distribution == "exp" ~ "Exponential",
                                  Distribution == "gamma" ~ "Gamma",
                                  Distribution == "lnorm" ~ "Log-normal",
                                  Distribution == "norm" ~ "Normal",
                                  Distribution == "unif" ~ "Uniform"), 
         Distribution = factor(Distribution), 
         PR = case_when(PRNAME == "Newfoundland and Labrador" ~ 10,
                        PRNAME == "Prince Edward Island" ~ 11,
                        PRNAME == "Nova Scotia" ~ 12,
                        PRNAME == "New Brunswick" ~ 13,
                        PRNAME == "Quebec" ~ 24,
                        PRNAME == "Ontario" ~ 35,
                        PRNAME == "Manitoba" ~ 46,
                        PRNAME == "Saskatchewan" ~ 47,
                        PRNAME == "Alberta" ~ 48,
                        PRNAME == "British Columbia" ~ 59,
                        PRNAME == "Yukon" ~ 60,
                        PRNAME == "Northwest Territories" ~ 61, 
                        PRNAME == "Nunavut" ~ 62), 
         PR = factor(PR), 
         CMACATEG = case_when(CMACATEG == "CMA/CA" ~ "Census Metropolitan Area or Census Agglomeration", 
                              TRUE ~ CMACATEG), 
         CMACATEG = factor(CMACATEG)
         )

pr_impedance_functions <- pr_impedance_functions %>%
  select("PR","PRNAME","CMACATEG",
         "PwMode_label","Distribution","est_1","est_2")
```

```{r}
# adding file to the package
usethis::use_data(pr_impedance_functions,
                  compress = "bzip2",
                  overwrite = TRUE)
```

# Land Use Tables 

## General Population 

```{r}
labour_force_CT_general <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/labour_force_CT_fixed.xlsx"), sheet = "General")

labour_force_CT_general <- labour_force_CT_general %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR))

usethis::use_data(labour_force_CT_general,
                  compress = "bzip2",
                  overwrite = TRUE)
```

```{r}
jobs_CT_general <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/jobs_CT_fixed.xlsx"), sheet = "General")

jobs_CT_general <- jobs_CT_general %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR))

usethis::use_data(jobs_CT_general,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Population by transportation mode

```{r}
labour_force_CT_mode <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/labour_force_CT_fixed.xlsx"), sheet = "Mode")

labour_force_CT_mode <- labour_force_CT_mode %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR), 
         PwMode = factor(PwMode), 
         ) %>% 
  rename(PwMode_label = PwMode)

usethis::use_data(labour_force_CT_mode,
                  compress = "bzip2",
                  overwrite = TRUE)
```
### Distributing non-commuters between the transportation modes 

```{r}
# Total commuters by CTUID and PwMode_label
commuters <- labour_force_CT_mode %>%
  filter(!PwMode_label == "Other or non-commuter") %>% 
  rename(commuters_mode = labour_force_rounded)

# commuter population by CTUID
commuters <- commuters %>%
             group_by(CTUID) %>% 
             mutate(all_commuters = sum(commuters_mode), 
                   perc_mode =  commuters_mode/all_commuters) 

# For cases when the CTUID there's only non_commuters respondents
general_perc_mode <- commuters %>%
  group_by(PCD, PwMode_label) %>%
  summarise(general_pop_mode = sum(commuters_mode)) %>%
  mutate(general_pop = sum(general_pop_mode),
         perc_pcd = general_pop_mode/general_pop)

commuters <- commuters %>% 
  left_join(general_perc_mode %>% dplyr::select(PCD, PwMode_label, perc_pcd), by = c("PCD", "PwMode_label")) 

# Total non commuters population by CTUID
non_commuters <- labour_force_CT_mode %>%
  filter(PwMode_label == "Other or non-commuter") %>% # Considering the category 'Other' as non-commuter
  rename(all_non_commuters = labour_force_rounded) 

# Allocating the non_commuters according to the share of each population mode
grid_exp <- expand.grid(CTUID = unique(labour_force_CT_mode$CTUID), 
                        PwMode_label = unique(commuters$PwMode_label))

commuters_exp <- grid_exp %>% 
  left_join(commuters %>% dplyr::select("CTUID", PwMode_label, 
                                        "commuters_mode", 
                                        "all_commuters",
                                        "perc_mode", 
                                        "perc_pcd"), 
            by = c("CTUID", "PwMode_label"))

labour_force_support <- commuters_exp %>% 
  left_join(non_commuters %>% dplyr::select("CTUID", 
                                        all_non_commuters), 
            by = c("CTUID"))

labour_force_support <- labour_force_support %>%
  mutate(perc_final = if_else(is.na(perc_mode), as.numeric(perc_pcd), perc_mode)) %>%
  group_by(CTUID) %>%
  mutate(perc_final = perc_final / sum(perc_final, na.rm = TRUE)) %>%
  ungroup() # making sure that the sum of mode percentages is equal to 1

labour_force_support <- labour_force_support %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

labour_force_support <- labour_force_support %>%
  mutate(est_labour_force = (all_non_commuters * perc_final) + commuters_mode) 
```



```{r selecting-var-labour-table}
estimated_labour_force_CT_mode <- labour_force_support %>% 
  dplyr::select(CTUID, PwMode_label, est_labour_force) %>% 
  left_join(labour_force_CT_mode %>% 
              dplyr::select("CTUID","CTNAME",
                            "PCD", "CMA",
                            "PR") %>% 
              distinct(),
            by = "CTUID") %>% 
  dplyr::select("CTUID","CTNAME","PCD","CMA",
                "PR","PwMode_label","est_labour_force")
```
```{r}
sum(estimated_labour_force_CT_mode$est_labour_force) == sum(labour_force_CT_mode$labour_force_rounded)

usethis::use_data(estimated_labour_force_CT_mode,
                  compress = "bzip2",
                  overwrite = TRUE)
```
## Population by education level 

```{r}
labour_force_CT_education <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/labour_force_CT_fixed.xlsx"), sheet = "Education")

labour_force_CT_education <- labour_force_CT_education %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR), 
         Education = case_when(Education == "College or certificate" ~ "College or apprenticeship certificate", 
                               Education == "High school or no certificate" ~ "High school or no certificate",
                               Education == "University certificate or higher" ~ "University certificate or higher"),
         Education = factor(Education), 
         )

usethis::use_data(labour_force_CT_education,
                  compress = "bzip2",
                  overwrite = TRUE)
```

```{r}
jobs_CT_occupation <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/jobs_CT_fixed.xlsx"), sheet = "Occupation")

jobs_CT_occupation <- jobs_CT_occupation %>% 
  rename(TEER = `Occupation level`) %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR), 
         TEER = case_when(TEER == "College or training" ~ "College or apprenticeship (2 - 3)", 
                               TEER == "High school or no training" ~ "High school or short-term experience (4 - 5)",
                               TEER == "Management or professional" ~ "Management or professional (0 - 1)"),
         TEER = factor(TEER), 
         )

usethis::use_data(jobs_CT_occupation,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Population by transportation mode and education level 

```{r}
labour_force_CT_education_mode <- read_excel(paste0(here::here(),"/data-ignore-inputs/release_RDC_08172026/labour_force_CT_fixed.xlsx"), sheet = "Education and Mode")

labour_force_CT_education_mode <- labour_force_CT_education_mode %>% 
  mutate(CTUID = factor(CTUID), 
         CTNAME = factor(CTNAME), 
         PCD = factor(PCD), 
         CMA = factor(CMA), 
         PR = factor(PR), 
                  Education = case_when(Education == "College or certificate" ~ "College or apprenticeship certificate", 
                               Education == "High school or no certificate" ~ "High school or no certificate",
                               Education == "University certificate or higher" ~ "University certificate or higher"),
         Education = factor(Education), 
         PwMode = factor(PwMode), 
         ) %>% 
  rename(PwMode_label = PwMode)

usethis::use_data(labour_force_CT_education_mode,
                  compress = "bzip2",
                  overwrite = TRUE)
```

### Distributing non-commuters between the transportation modes and education level

```{r}
# Total commuters by CTUID, Education, and PwMode_label
commuters <- labour_force_CT_education_mode %>%
  filter(!PwMode_label == "Other or non-commuter") %>%
  rename(commuters_mode = labour_force_rounded)

# share of each mode within CTUID - Education
commuters <- commuters %>%
  group_by(CTUID, Education) %>%
  mutate(all_commuters_ct_edu = sum(commuters_mode),
         perc_mode_ct_edu = commuters_mode / all_commuters_ct_edu) %>%
  ungroup()

# share of each mode within PCD - Education 
general_perc_mode_pcd_edu <- commuters %>%
  group_by(PCD, Education, PwMode_label) %>%
  summarise(commuters_mode_sum = sum(commuters_mode), .groups = "drop") %>%
  group_by(PCD, Education) %>%
  mutate(all_commuters_pcd_edu = sum(commuters_mode_sum),
         perc_mode_pcd_edu = commuters_mode_sum / all_commuters_pcd_edu) %>%
  ungroup()

# share of each mode within CTUID, ignoring Education
general_perc_mode_ct <- commuters %>%
  group_by(CTUID, PwMode_label) %>%
  summarise(commuters_mode_sum = sum(commuters_mode), .groups = "drop") %>%
  group_by(CTUID) %>%
  mutate(all_commuters_ct = sum(commuters_mode_sum),
         perc_mode_ct = commuters_mode_sum / all_commuters_ct) %>%
  ungroup()

# share of each mode within PCD, ignoring Education 
general_perc_mode_pcd <- commuters %>%
  group_by(PCD, PwMode_label) %>%
  summarise(commuters_mode_sum = sum(commuters_mode), .groups = "drop") %>%
  group_by(PCD) %>%
  mutate(all_commuters_pcd = sum(commuters_mode_sum),
         perc_mode_pcd = commuters_mode_sum / all_commuters_pcd) %>%
  ungroup()

# non-commuters population by CTUID and Education
non_commuters <- labour_force_CT_education_mode %>%
  filter(PwMode_label == "Other or non-commuter") %>%
  rename(all_non_commuters = labour_force_rounded) %>%
  dplyr::select(CTUID, Education, all_non_commuters)

ct_pcd_lookup <- labour_force_CT_education_mode %>% distinct(CTUID, PCD)
ct_edu_combos <- labour_force_CT_education_mode %>% distinct(CTUID, Education)

grid_exp <- ct_edu_combos %>%
  tidyr::crossing(PwMode_label = unique(commuters$PwMode_label)) %>%
  left_join(ct_pcd_lookup, by = "CTUID")

labour_force_support <- grid_exp %>%
  left_join(commuters %>% dplyr::select(CTUID, Education, 
                                        PwMode_label,
                                         commuters_mode,perc_mode_ct_edu),
            by = c("CTUID", "Education","PwMode_label")) %>%
  left_join(general_perc_mode_pcd_edu %>%
              dplyr::select(PCD, Education,PwMode_label,perc_mode_pcd_edu),
            by = c("PCD", "Education","PwMode_label")) %>%
  left_join(general_perc_mode_ct %>%
              dplyr::select(CTUID, PwMode_label, perc_mode_ct),
            by = c("CTUID","PwMode_label")) %>%
  left_join(general_perc_mode_pcd %>%
              dplyr::select(PCD, PwMode_label, perc_mode_pcd),
            by = c("PCD","PwMode_label")) %>%
  left_join(non_commuters, by = c("CTUID","Education"))

labour_force_support <- labour_force_support %>%
  mutate(perc_final = case_when(!is.na(perc_mode_ct_edu) ~ perc_mode_ct_edu,
                                !is.na(perc_mode_pcd_edu) ~ perc_mode_pcd_edu,
                                !is.na(perc_mode_ct) ~ perc_mode_ct,
                                TRUE ~ perc_mode_pcd)) %>%
  group_by(CTUID, Education)%>%
  mutate(perc_final = perc_final/sum(perc_final, na.rm=TRUE))%>%
  ungroup()

labour_force_support <- labour_force_support %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

labour_force_support <- labour_force_support %>%
  mutate(est_labour_force = (all_non_commuters*perc_final) + commuters_mode)
```

```{r}
estimated_labour_force_CT_education_mode <- labour_force_support %>%
  dplyr::select(CTUID, Education, PwMode_label, est_labour_force) %>%
  left_join(labour_force_CT_education_mode %>%
              dplyr::select(CTUID, CTNAME, PCD, CMA, PR) %>%
              distinct(),
            by = "CTUID") %>%
  dplyr::select(CTUID, CTNAME, PCD, CMA, PR, Education, PwMode_label, est_labour_force)
```

```{r}
sum(estimated_labour_force_CT_education_mode$est_labour_force) == sum(labour_force_CT_education_mode$labour_force_rounded)

usethis::use_data(estimated_labour_force_CT_education_mode,
                  compress = "bzip2",
                  overwrite = TRUE)
```
# Maximum Travel Times

## Census Division Level

```{r}
# reading file
pcd_max_travel_times <- read_csv(glue::glue(here::here(), 
                                            "/data-ignore-inputs/pcd_max_travel_times.csv"))

pcd_max_travel_times <- pcd_max_travel_times %>%
  rename("PwMode_label" = PwMode, Maximum_travel_time = Max_time_value) %>%
  mutate(PwMode_label = case_when(PwMode_label == "Car/motor" ~ "Car-motorized", 
                                  PwMode_label == "Transit" ~ "Public transit", 
                                  TRUE ~ PwMode_label),
         PwMode_label = factor(PwMode_label),
         PCD = factor(PCD))

pcd_max_travel_times <- pcd_max_travel_times %>% 
  left_join(pcd_keys, by = "PCD") %>% 
  dplyr::select("PCD","PCDNAME","PR","PRNAME","PwMode_label","Maximum_travel_time")
```

```{r}
# adding file to the package
usethis::use_data(pcd_max_travel_times,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Census Metropolitan Area Level

```{r}
# reading file
cma_max_travel_times <-read_csv(glue::glue(here::here(), 
                                           "/data-ignore-inputs/cma_max_travel_times.csv"))

cma_max_travel_times <- cma_max_travel_times %>% 
  filter(CMA > 0, CMA < 996) %>% 
  rename("PwMode_label" = PwMode, Maximum_travel_time = Max_time_value) %>%
  mutate(PwMode_label = case_when(PwMode_label == "Car/motor" ~ "Car-motorized", 
                                  PwMode_label == "Transit" ~ "Public transit", 
                                  TRUE ~ PwMode_label),
         PwMode_label = factor(PwMode_label),
         CMA = factor(sprintf("%03d", as.numeric(CMA)))
         ) 

cma_max_travel_times <- cma_max_travel_times %>% 
  left_join(cma_keys, by = "CMA") %>% 
  dplyr::select("CMA","CMANAME","PR","PRNAME","PwMode_label","Maximum_travel_time")
```

```{r}
# adding file to the package
usethis::use_data(cma_max_travel_times,
                  compress = "bzip2",
                  overwrite = TRUE)
```


<!--

Update documentation:
```{r}
devtools::document()
```

Verifying if the package has any problem:
```{r}
devtools::check()
```

Building the package:
```{r}
devtools::build()
```

Updating github webpage:
```{r}
usethis::use_pkgdown_github_pages()
```
-->

