---
title: "Adding_Datasets"
output: html_document
---

Load libraries:
```{r}
library('sf')
library('tidyverse')
library('usethis')
library('devtools')
```

# Spatial Data

## Census divisions:
```{r}
# reading file
census_divisions_ca21 <- sf::read_sf("C:/Bruno/packages/0_support_packages/commute_CA/lcd_buffer.shp")
```

```{r}
# adding file to the package
usethis::use_data(census_divisions_ca21,
                  compress = "bzip2",
                  overwrite = TRUE)
```

## Census Tracts
```{r}
# reading file
census_tracts_ca21 <- sf::read_sf("C:/Bruno/packages/0_support_packages/commute_CA/census_tracts_ca21.shp")
```

```{r}
# adding file to the package
usethis::use_data(census_tracts_ca21,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## Dissemination Areas
```{r}
# reading file
dissemination_areas_ca21 <- sf::read_sf("C:/Bruno/packages/0_support_packages/commute_CA/dissemination_areas_ca21_str.shp")
```

```{r}
# adding file to the package
usethis::use_data(dissemination_areas_ca21,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```
# Tabular data

## Census Test Data:
```{r}
# reading file
census_test_data <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/census_test_v3.csv")
```

```{r}
# adding file to the package
usethis::use_data(census_test_data,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## CMA Impedance Functions
```{r}
# reading file
cma_impedance_functions <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/cma_impedance_functions.csv")
```

```{r}
# adding file to the package
usethis::use_data(cma_impedance_functions,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```


## PCD Impedance Functions
```{r}
# reading file
pcd_impedance_functions <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/pcd_impedance_functions.csv")
```

```{r}
# adding file to the package
usethis::use_data(pcd_impedance_functions,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## PR Impedance Functions
```{r}
# reading file
pr_impedance_functions <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/pr_impedance_functions.csv")
```

```{r}
# adding file to the package
usethis::use_data(pr_impedance_functions,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## Land Use by mode 
```{r}
# reading file
land_use_CT_mode <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/land_use_CT_mode.csv")
```

```{r}
# adding file to the package
usethis::use_data(land_use_CT_mode,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

## Land Use general
```{r}
# reading file
land_use_CT_general <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/land_use_CT_general.csv")
```

```{r}
# adding file to the package
usethis::use_data(land_use_CT_general,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

# Maximum Travel Times PCD

```{r}
# reading file
pcd_max_travel_times <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/pcd_max_travel_times.csv")
```

```{r}
# adding file to the package
usethis::use_data(pcd_max_travel_times,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```
# Maximum Travel Times CMA

```{r}
# reading file
cma_max_travel_times <- read_csv("C:/Bruno/packages/0_support_packages/commute_CA/cma_max_travel_times.csv")
```

```{r}
# adding file to the package
usethis::use_data(cma_max_travel_times,
                   compress = "xz",
                  version = 3,
                  overwrite = TRUE)
```

Update documentation:
```{r}
devtools::document()
```

Verifying if the package has any problem:
```{r}
devtools::check()
```

Building the package:
```{r}
devtools::build()
```

Updating github webpage:
```{r}
usethis::use_pkgdown_github_pages()
```

