---
title: "Defining equity standards based on the relationship between job accessibility and employment outcomes"
runningheader: "a methodology for Canadian regions based on the 2021 Census of Population" # only for pdf output
subtitle: "a methodology for Canadian regions based on the 2021 Census of Population" # only for html output
author: "Bruno Santos & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r clean-workspace, include=FALSE}
# cleaning objects from the workspace 
rm(list = ls())
```

```{r setup, include=FALSE}
# layout configuration 
library(tufte)
library(knitr)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

This Rmarkdown file is part of the
[**CommuteCA**](https://github.com/dias-bruno/CommuteCA) package. This
package was created in conjunction with the office of the [*Research
Data Center* at *McMaster University*](https://rdc.mcmaster.ca/), the
[*Sherman Centre for Digital Scholarship*](https://scds.ca/) and the
[*Mobilizing Justice*](https://mobilizingjustice.ca/)[^1].

[^1]: For this demonstration, we will use the case of the city of
    Toronto. If you want to use the other options, the original data
    from the Census in an RDC office or the test data for all locations
    in Canada, *update* the address in the chunk.

The [**CommuteCA**](https://github.com/dias-bruno/CommuteCA) R package
was created to develop standardized methods for transport analysis in
research, particularly for analysis using the [*2021 Census of
Population*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm)
from Statistics Canada. We focused our efforts on the [*Commuting
Reference
Guide*](https://www12.statcan.gc.ca/census-recensement/2021/ref/98-500/011/98-500-x2021011-eng.cfm),
which provides valuable variables and information on commuting for the
Canadian population aged 15 and older living in private households.

# Section objectives

In the previous section, we developed spatial filters to address the likely endogeneity between accessibility measures and individual outcomes. Building on this, we now develop regression models to examine the relationship between job accessibility and employment outcomes, with the goal of identifying key accessibility thresholds associated with improved results in employment status and job seeking engagement.

We first model the relationship between each dependent variable (employment outcome) and a set of predictor variables, which include job accessibility indicators and socioeconomic characteristics. These models estimate the probability of an outcome occurring, conditional on a linear combination of the predictors. All models are estimated separately for disadvantaged and non-disadvantaged populations, as defined earlier in Section 6.

Following this modelling stage, we identify equity-based accessibility standards: key thresholds of job accessibility for disadvantaged populations that correspond to more favourable employment outcomes.

## Suggested Readings

 -  Bastiaanssen, J., Johnson, D., & Lucas, K. (2025). *The relationship between individual employment probabilities and accessibility to matching jobs: A study of the Netherlands*. Transportation Research Part A: Policy and Practice, 193, 104398. 
 -  Cunningham, S. (2021). *Causal inference: The mixtape*. Yale
    university press.

# Let's code!

Load the packages:

```{r load-packages}
library(dplyr) # A Grammar of Data Manipulation
library(here) # Enables easy file referencing in project-oriented workflows
library(survey) # Tools for analyzing complex survey samples with weights, strata, and clustering
library(svydiags)
library(tidyr) # Tools for tidying data: reshaping, pivoting, separating, and uniting columns
library(VGAM) # Vector generalized linear and additive models, including multinomial and ordinal models
library(stringr) # Consistent, simple functions for working with strings
library(MASS) # Functions and datasets from *Modern Applied Statistics with S*, including statistical models
library(car) # Applied regression tools, including diagnostics, recoding, and ANOVA utilities
library(ggplot2) # Create Elegant Data Visualizations Using the Grammar of Graphics
library(sf) # Support for simple features: a standardized way to encode spatial vector data
library(openxlsx) # Read, write, and format Excel files
```


```{r}
# install order for the packages 
packages <- c("VGAM", "cli",  "generics",     
"glue",          "magrittr",      "R6",           
"rlang",         "utf8",          "pkgconfig",    
"withr",         "rprojroot",     "lattice",      
"numDeriv",      "Rcpp",          "DBI",          
"cpp11",         "stringi",       "MASS",         
"carData",       "abind",         "Formula",      
 "nnet",          "boot",          "nloptr",       
 "SparseM",       "farver",        "labeling",     
"RColorBrewer",  "viridisLite",   "backports",    
 "Deriv",         "microbenchmark","rbibutils",    
"colorspace",    "fracdiff",      "timeDate",     
"S7",            "KernSmooth",    "wk",           
"proxy",         "zip",           "lifecycle",    
 "here",          "Matrix",        "nlme",         
 "zoo",           "minqa",         "RcppArmadillo",
 "RcppEigen",     "units",         "mitools",      
 "isoband",       "class",         "Rdpack",       
 "s2",            "openxlsx",      "vctrs",        
 "scales",        "gtable",        "survival",     
 "MatrixModels",  "mgcv",          "urca",         
 "lmtest",        "e1071",         "reformulas",   
 "stringr",       "pillar",        "tidyselect",   
 "purrr",         "ggplot2",       "survey",       
 "quantreg",      "classInt",      "lme4",         
 "tibble",        "cowplot",       "forecast",     
 "svydiags",      "sf",            "dplyr",        
 "tidyr",         "broom",         "modelr",       
 "doBy",          "pbkrtest",      "car")

zip_folder <- "Z:/VRDC-PROJ-10840/Santos_10840/santosb/MCM-BRUNO-10840/"

zip_files <- paste0(zip_folder,packages,".zip")

zip_files
```

## Data and Folders

Reading census data and creating a R data frame[^2]:

[^2]: For this demonstration, we will use the case of the city of
    Toronto. If you want to use the other options, the original data
    from the Census in an RDC office or the test data for all locations
    in Canada, *update* the address in the chunk.

```{r}
# For the original Census data set
files_address <- paste0(here(),"/data-ignore-inputs/census_test_GTAH.csv") # insert-the-RDC-address!!!
census <- read.csv(files_address, header = TRUE)
census_backup <- census
```

If the original Census dataset available in the RDC is not in .csv
(comma-separated values) format but is instead provided in other formats
such as SPSS, SAS, or SAS Data, you can use the foreign package (a
built-in R library) to import it:

```{r}
# library(foreign)
# foreign::read.dta(files_address) # For Stata
# foreign::read.spss(files_address) # For SPSS
# There are many other options! You can search for this library in the 'Packages' window and explore additional functions for reading your file. 
```

Or if you have installed *CommuteCA*, you can access the *test* data:

```{r load-census-data}
# data("census_test_toronto")
# census <- census_test_toronto
```

| ⚠️**NOTE:** If the code above did not run correctly, you probably are experiencing a file address error. Try to identify the correct address and update the chunk named `census-file-address` to continue.


Setting the folder with the accessibility measures and spatial filter:

```{r}
study_area_label <- "GTAH"
spatial_unit <- "CT"
spatial_unit_identifier <- "CTUID"

study_area_folder <- paste0(here::here(),"/data-raw/output/",
study_area_label,"/")
study_area_folder
```


```{r read-accessibility}
# Folder  
measures_folder <- paste0(study_area_folder,"/",spatial_unit,"/spatial-filter-measures/")

# Read file
SA_mode <- read.csv(paste0(measures_folder, "spatial_filters.csv"), colClasses = c("character", "character", 
                              "numeric","numeric",
                              "numeric","numeric", 
                              "numeric","numeric")) %>% 
  mutate(PwMode_label = case_when(PwMode_label == "Car-motor" ~ "Car_motor",
    TRUE ~ PwMode_label ))
```

```{r}
SA_mode_wide <- SA_mode %>%
  rename(SA = SA_im, 
         HT = HT_im,
         SA_pp = SA_pp_im) %>%
  pivot_wider(id_cols = spatial_unit_identifier,
              names_from = PwMode_label, 
              values_from = c(SA, SA_pp, HT, Sf_dist_SA, Sf_dist_SA_pp, Sf_dist_HT),
              names_sep = "_") 

SA_mode_wide
```

Reading the dissemination area geometries:

```{r}
directory_spatial_files <- paste0(study_area_folder,"spatial-files/") # Update this address if necessary
dissemination_areas <- st_read(paste0(directory_spatial_files, "dissemination_areas.gpkg"))

dissemination_areas <- dissemination_areas %>% 
  mutate(DAUID = as.numeric(DAUID))
```

Reading the census tract geometries: 

```{r}
census_tracts <- st_read(paste0(directory_spatial_files, "census_tracts.gpkg"))
```

Reading the census division geometries:

```{r}
census_subdivisions <- st_read(paste0(directory_spatial_files, "census_subdivisions.gpkg"))
```

Setting folders to export files:

```{r directory-export}
diretorio_export <- paste0(study_area_folder, spatial_unit,  "/standards-measures/") # Update the address to export the data, if necessary

if(!dir.exists(diretorio_export)){
  dir.create(diretorio_export, recursive = TRUE)}

diretorio_export_images <- paste0(study_area_folder, spatial_unit,"/standards-figures/") # Update the address to export the data, if necessary

if(!dir.exists(diretorio_export_images)){
  dir.create(diretorio_export_images, recursive = TRUE)}
```

## Preprocessing files

We will filter the census data frame by administrative (province
and/or census division) and/or statistical (census metropolitan areas
and census agglomerations)[^3]. The chunk below shows how to make this
procedure:

[^3]: As said before, we will perform our analysis for the city of
    Toronto. If you want to select a specific area to work, uncomment
    the code above that applies to your case and select the apropriate
    code of your interest unit. Please, check the dictionary to have
    more informations about the [provinces
    code](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/tab/index-eng.cfm?ID=t1_8),
    [census
    divisions](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo008),
    and [census metropolitan
    areas](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo009)

```{r filter-boundaries}
pcd_codes <- c("3518", # Durham
          "3519", # York
          "3520" ,  # Toronto
          "3521", # Peel
          "3524", # Halton
          "3525" # Hamilton
          ) 

# Filtering by census division
 census_filtered <- census %>%
                    filter(PCD %in% pcd_codes) # Only select respondents who live in  Toronto
```

According to the census code book, the variable 'PwMode' has the
following possible values:

-   -3: Not applicable.
-   1: Car, truck or van - as a driver.
-   2: Car, truck or van - as a passenger.
-   3: Bus.
-   4: Subway or elevated rail.
-   5: Light rail, streetcar or commuter train.
-   6: Passenger ferry.
-   7: Walked.
-   8: Bicycle.
-   9: Motorcycle, scooter or moped.
-   10: Other method.

We'll rename the travel modes to facilitate the readability of the data.
Additionally, we'll remove from our analysis travel modes signed as
'Other methods':

According to the Census 2021, an 'Employed' person refers to those who,
during the reference period, had a labour force status of
'employed.'[^4]. In addition, the labour force is made up of the
employed population and the unemployed population. This variable can
assume the values:

[^4]: Those who, during the reference period: (a) Did any work at all at
    a job or business, that is, paid work in the context of an
    employer-employee relationship, or self-employment. This also
    includes persons who did unpaid family work, which is defined as
    unpaid work contributing directly to the operation of a farm,
    business or professional practice owned and operated by a related
    member of the same household; or (b) Had a job but were not at work
    due to factors such as their own illness or disability, personal or
    family responsibilities, vacation or a labour dispute. This category
    excludes persons not at work because they were on layoff or between
    casual jobs, and those who did not then have a job (even if they had
    a job to start at a future date).

-   -3: Not Applicable, \< 15 years
-   1: In Labour Force, Employed
-   2: In Labour Force, Unemployed
-   3: Not in Labour Force

```{r lf-selection}
census_vetting <- census_filtered  %>%
  dplyr::select("Frame_ID", "PRCDDA", "PR", "CMA",  "PCD", "POP_CNTR_IND","CompW1",
                
                "LBR_FORC", "CfInc_AT", "CFCNT_PP", "CF_PnCF_NumEarners", "PWDA",
                "PWPR", "PWCD", "PWCMA", "PWDUR", "PWDist", "PwMode", "AGE",
                #"MRNFY", 
                "GENDER_BINARY", "LoLICOA", "LoLIMA", "ACTLMFLG", 
                "CFAMSTSIMPLE", "IMMCAT5", "PerImmA", "DVisMin", 
                "HCDD", "CFKID0T5", "Weeks", "LOKWK","Mob5", "GovtI_SocAssist", 
                #"EMPLOYED","LF71"
                )

census_filtered <- census_filtered %>% 
  filter(LBR_FORC == 1 | LBR_FORC == 2) %>%
  dplyr::select(names(census_vetting))
```

According to the Census 2021, an 'Employed' person refers to those who,
during the reference period, had a labour force status of
'employed.'[^5]. In addition, the labour force is made up of the
employed population and the unemployed population. This variable can
assume the values:

[^5]: Those who, during the reference period: (a) Did any work at all at
    a job or business, that is, paid work in the context of an
    employer-employee relationship, or self-employment. This also
    includes persons who did unpaid family work, which is defined as
    unpaid work contributing directly to the operation of a farm,
    business or professional practice owned and operated by a related
    member of the same household; or (b) Had a job but were not at work
    due to factors such as their own illness or disability, personal or
    family responsibilities, vacation or a labour dispute. This category
    excludes persons not at work because they were on layoff or between
    casual jobs, and those who did not then have a job (even if they had
    a job to start at a future date).

-   -3: Not Applicable, \< 15 years (already previously excluded from
    our dataset)
-   1: In Labour Force, Employed
-   2: In Labour Force, Unemployed
-   3: Not in Labour Force (already previously excluded from our
    dataset)

According to the census code book, the variable 'GENDER_BINARY' has the
following possible values:[^6]

[^6]: Given that the non-binary population is small, data aggregation to
    a two-category gender variable is sometimes necessary to protect the
    confidentiality of responses provided. In these cases, individuals
    in the category “non-binary persons” are distributed into the other
    two gender categories and are denoted by the “+” symbol.

-   1: Woman+.
-   2: Man+.

According to the census code book, the variable 'GENDER_BINARY' has the
following possible values:

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Concept not applicable (territories or reserves) - not in low
    income
-   4: Concept not applicable (territories or reserves) - in low income

According to the census code book, the variable 'LoLICOA' has the
following possible values:[\^]

[\^] Refers to the income situation of a person in relation to the
Low-income cut-offs, after-tax (LICO-AT). Persons with income that is
below this low-income line are considered to be in low income.

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Concept not applicable (territories or reserves) - not in low
    income
-   4: Concept not applicable (territories or reserves) - in low income

According to the census code book, the variable 'LoLIMA' has the
following possible values:[\^]

[\^] Refers to the income situation of a person in relation to the
Low-income measure, after-tax (LIM-AT). Persons with income that is
below this low-income line are considered to be in low income.

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Not in low income (territories or reserves)
-   4: In low income (territories or reserves)
-   5: Concept not applicable (collective households) - not in low
    income
-   6: Concept not applicable (collective households) - in low income

We will use the variable *ACTLMFLG - Activities of daily living: Flag*
as an indicative of disability, because the Census of Population does
not have a variable regarding disability. The question on activities of
daily living refers to difficulties a person may have doing certain
activities as a result of physical, cognitive, mental, or other
health-related conditions or problems. We will select all those
respondents who affirm to have frequent (always) difficulties to develop
regular activities. According to the census code book, the variable
'ACTLMFLG' has the following possible values:

-   -2: Not stated
-   1: Yes, always
-   2: Yes, often
-   3: Yes, sometimes
-   4: No

According to the census code book, the variable 'CFAMSTSIMPLE' has the
following possible values:

-   1: Married spouse or common-law partner
-   2: Parent in a one-parent family
-   3: Child
-   4: Person not in a census family

According to the census code book, the variable 'IMMCAT5' has the
following possible values:

-   1: Non-immigrants
-   2: Non-permanent residents3 Immigrants admitted before 1980
-   100010: Economic immigrants - Principal applicants
-   100020: Economic immigrants - Secondary applicants
-   200000: Immigrants sponsored by family
-   300000: Refugees
-   400000: Other immigrants

According to the census code book, "Visible minority" refers to whether
a person is a visible minority as defined by the Employment Equity Act.
The Employment Equity Act defines visible minorities as "persons, other
than Aboriginal peoples, who are non-Caucasian in race or non-white in
colour." The visible minority population consists mainly of the
following groups: South Asian, Chinese, Black, Filipino, Arab, Latin
American, Southeast Asian, West Asian, Korean and Japanese. The variable
'DVisMin' has the following possible values:

-   1: Other (not a visible minority)
-   2: South Asian
-   3: Chinese
-   4: Black
-   5: Filipino
-   6: Arab
-   7: Latin American
-   8: Southeast Asian
-   9: West Asian
-   10: Korean
-   11: Japanese
-   12: Visible minority, n.i.e.
-   13: Multiple visible minorities
-   14: Indigenous peoples

In the next chunk, we will rename the travel modes to facilitate the
readability of the data.

```{r labelling-variables}
census_filtered_labeled <- census_filtered %>%
                   mutate(PwMode_label = 
                            case_when(PwMode > 0 & PwMode <= 2 ~ "Car_motor",
                                      PwMode == 9 ~ "Car_motor",
                                      PwMode >= 3 & PwMode <= 6  ~ "Transit",
                                      PwMode == 7  ~ "Walk",
                                      PwMode == 8  ~ "Bike", 
                                      PwMode == 10 ~ "Other",
                                      PwMode == -3 ~ "Not applicable"),
                          PwMode_label = factor(PwMode_label, levels = c("Bike", "Walk", "Car_motor", "Transit","Not applicable","Other")),
                          # Walk 
                          Walk = 
                            case_when(PwMode == 7  ~ "Yes",
                                      TRUE ~ "No"),
                          Walk = factor(Walk, levels = c("No", "Yes")),
                     # Bike 
                          Bike = 
                            case_when(PwMode == 8 ~ "Yes",
                                      TRUE ~ "No"),
                          Bike = factor(Bike, levels = c("No", "Yes")),
                     # Active 
                          Transit = 
                            case_when(PwMode_label == "Transit"  ~ "Yes",
                                      TRUE ~ "No"),
                          Transit = factor(Transit, levels = c("No", "Yes")),
                     # Active 
                          Active = 
                            case_when(PwMode == 7  ~ "Yes",
                                      PwMode == 8  ~ "Yes", 
                                      TRUE ~ "No"),
                          Active = factor(Active, levels = c("No", "Yes")),
                     # Labour force
                     LBR_FORC_label = 
                            case_when(LBR_FORC == 1 ~ "Employed",
                                      LBR_FORC == 2 ~ "Unemployed"),
                          LBR_FORC_label = factor(LBR_FORC_label), 
                     # Gender 
                          Gender = 
                            case_when(GENDER_BINARY == 1 ~ "Woman+",
                                      GENDER_BINARY == 2 ~ "Man+"),
                          Gender = factor(Gender, levels = c("Man+", "Woman+")),
                     # Low income population - LoLICOA
                     LoLICOA_label = 
                       case_when(LoLICOA == -3 ~ "Not in low income", 
                                 LoLICOA == 1 ~ "Not in low income",
                                 LoLICOA == 2 ~ "In low income",
                                 LoLICOA == 3 ~ "Not in low income",
                                 LoLICOA == 4 ~ "In low income"),
                     LoLICOA_label = factor(LoLICOA_label), 
                    LoLIMA_label = 
                       case_when(LoLIMA == -3 ~ "Not in low income", 
                                 LoLIMA == 1 ~ "Not in low income",
                                 LoLIMA == 2 ~ "In low income",
                                 LoLIMA == 3 ~ "Not in low income",
                                 LoLIMA == 4 ~ "In low income",
                                 LoLIMA == 5 ~ "Not in low income",
                                 LoLIMA == 6 ~ "In low income"),
                     LoLIMA_label = factor(LoLIMA_label), 
                    Low_Income_label = 
                      case_when(LoLIMA_label == "In low income" | LoLICOA_label == "In low income" ~ "In low income",
                                TRUE ~ "Not in low income"), 
                    Low_Income_label = factor(Low_Income_label, levels = c("Not in low income", "In low income")), 
                    # Disability
                    Disability = 
                       case_when(ACTLMFLG == -2 ~ "No", 
                                 ACTLMFLG == 1 ~ "Yes",
                                 ACTLMFLG == 2 ~ "No",
                                 ACTLMFLG == 3 ~ "No",
                                 ACTLMFLG == 4 ~ "No"),
                    Disability = factor(Disability, levels = c("No", "Yes")), 
                    # Single-parent family
                    Single_parent = 
                       case_when(CFAMSTSIMPLE == 1 ~ "No",
                                 CFAMSTSIMPLE == 2 ~ "Yes",
                                 CFAMSTSIMPLE == 3 ~ "No",
                                 CFAMSTSIMPLE == 4 ~ "No"),
                     Single_parent = factor(Single_parent, levels = c("No", "Yes")), 
                    # Citizenship
                    Temporary_or_refugee = 
                       case_when(IMMCAT5 == 2 ~ "Yes",
                         IMMCAT5 == 300000 ~ "Yes",
                                 TRUE ~ "No"),
                     Temporary_or_refugee = factor(Temporary_or_refugee, levels = c("No","Yes")), 
                    # New immigrant
                    PerImmA_label = 
                       case_when(PerImmA == 10 ~ "New immigrant",
                                 PerImmA != 10 ~ "Not a new immigrant"),
                     PerImmA_label = factor(PerImmA_label),
                    # Visible Minority (binary)
                    Visible_minority = 
                       case_when(DVisMin == 1 ~ "No",
                                 DVisMin != 1 ~ "Yes"),
                     Visible_minority = factor(Visible_minority, 
                                                   levels = c("No","Yes")), 
                    # Visible Minority (full-group)
                    DVisMin_label = 
                      case_when(DVisMin == 1  ~ "Other (not a visible minority)",
                                DVisMin == 2  ~ "South Asian",
                                DVisMin == 3  ~ "Chinese",
                                DVisMin == 4  ~ "Black",
                                DVisMin == 5  ~ "Filipino",
                                DVisMin == 6  ~ "Arab",
                                DVisMin == 7  ~ "Latin American",
                                DVisMin == 8  ~ "Southeast Asian",
                                DVisMin == 9  ~ "West Asian",
                                DVisMin == 10 ~ "Korean",
                                DVisMin == 11 ~ "Japanese",
                                DVisMin == 12 ~ "Visible minority, n.i.e.",
                                DVisMin == 13 ~ "Multiple visible minorities",
                                DVisMin == 14 ~ "Indigenous peoples"),
                    DVisMin_label = factor(DVisMin_label),
                    # AGE
                    AGE_label = 
                      case_when(AGE <= 24 ~ "Youth",
                                TRUE  ~ "Adult"),
                    AGE_label = factor(AGE_label), 
                    # Number of Weeks worked in 2020 
                    Weeks = 
                      case_when(Weeks < 0 ~ 0, 
                                TRUE ~ Weeks), 
                    # Low educated 
                    HCDD_label =
                      case_when(HCDD == 1 ~ "Low educated", 
                                TRUE ~ "High-school diploma"),
                    HCDD_label = factor(HCDD_label, levels = c("High-school diploma", 
                                                               "Low educated")),
                    # Education
                    Education = case_when(HCDD < 3  ~ "No certificate or high school only",
                      HCDD %in% 3:8 ~ "Post-secondary non-degree",
                      HCDD %in% 9:13 ~ "University degree"),
                    Education = factor(Education, levels = c("No certificate or high school only", "Post-secondary non-degree", "University degree")),
                    # Looked for paid work in the previous month 
                    LOKWK_label = 
                      case_when(LOKWK == -5 ~ "Employed",
                                LOKWK == -3 ~ "Not applicable",
                                LOKWK == 1 ~  "No",
                                LOKWK == 2 ~  "Yes (for full-time work)",
                                LOKWK == 3 ~ "Yes (for part-time work)"),
                    LOKWK_label = factor(LOKWK_label, levels = c("Yes (for full-time work)",
                                                     "Yes (for part-time work)", 
                                                     "No", 
                                                     "Employed", 
                                                     "Not applicable")), 
                    # Population centre
                    Pop_centre =
                      case_when(POP_CNTR_IND == 1 ~ "Rural",
                                POP_CNTR_IND == 2 ~ "Small",
                                POP_CNTR_IND == 3 ~  "Medium",
                                POP_CNTR_IND == 4 ~  "Large"),
                    Pop_centre = factor(Pop_centre, levels = c("Rural", "Small", "Medium", "Large")), 
                    Children0to5 = case_when(is.na(CFKID0T5) ~ 0,
                                             TRUE ~ CFKID0T5)
                    
                    )
```

Creating dummy variables for Woman+, Low educated, Disability, Racialized and Single Parents:

```{r}
census_filtered_labeled <- census_filtered_labeled %>%
                    mutate(woman_dummy = if_else(Gender == "Woman+", 1L, 0L),
                           low_educated_dummy = if_else(HCDD_label == "Low educated", 1L, 0L),
                           disability_dummy = if_else(Disability == "Yes", 1L, 0L),
                           racialized_dummy = if_else(Visible_minority == "Yes", 1L, 0L),
                           single_parent_dummy = if_else(Single_parent == "yes", 1L, 0L),
                           employed_dummy = if_else(LBR_FORC_label == "Employed", 1L, 0L))
```

Creating neighborhood variables: unemployed rate, percentage of car commuters, and DA population density.

```{r}
# Unemployment rate by DA
unemp_rate <- census_filtered_labeled %>%
  group_by(PRCDDA) %>%
  summarise(employed_w = sum(CompW1[LBR_FORC == 1], na.rm = TRUE),
    unemployed_w = sum(CompW1[LBR_FORC == 2], na.rm = TRUE),
    .groups = "drop") %>%
  mutate(unemployment_rate = (unemployed_w / (employed_w + unemployed_w)) * 100)

# Car commuters rate 
car_rate <- census_filtered_labeled %>%
  filter(PwMode_label != "Not applicable") %>%
  group_by(PRCDDA) %>%
  summarise(car_w   = sum(CompW1[PwMode_label == "Car_motor"], na.rm = TRUE),
    total_w = sum(CompW1, na.rm = TRUE), 
    .groups = "drop") %>%
  mutate(car_share = (car_w / total_w) * 100)

# DA Population Density

pop_density_da <- dissemination_areas %>%
  mutate(Pop_density = Population / LANDAREA) %>% 
  dplyr::select(DAUID, Pop_density) %>% 
  st_drop_geometry()

census_filtered_labeled <- census_filtered_labeled %>%
  left_join(pop_density_da,
            by = c("PRCDDA" = "DAUID")) %>%
  left_join(unemp_rate[,c("PRCDDA","unemployment_rate")],
            by = "PRCDDA") %>%
  left_join(car_rate[,c("PRCDDA","car_share")],
            by = "PRCDDA") %>% 
  mutate(     car_share = case_when(is.na(car_share) ~ 0, 
                               TRUE ~ car_share)
  )
```

View summary statistics from the data frame:

```{r census-summary}
summary(census_filtered_labeled)
```

Now, we create a table called respondents that will serve as the base for building the regression models. To better approximate a causal relationship between accessibility and employment outcomes, we restrict the sample according to the following criteria:

- We include only respondents who have lived at the same address for at least five years, in order to reduce residence self-selection bias.
- Because the Census does not contain information on household car ownership, and because car ownership is endogenous to employment outcomes, we select only individuals living in areas where no more than 50% of commuters travel by car.
- We exclude individuals who did not work the full year due to illness or disability; pregnancy, maternity, or parental leave; full-time studies or training; caregiving responsibilities; or retirement. This restriction focuses the analysis on respondents for whom accessibility is more likely to influence employment outcomes.

```{r respondents}
keys <- read.csv(paste0(here(), "/data-raw/input/census_spatial_keys.csv"), 
                 colClasses = c("character", "character", 
                              "character","character",
                              "character","character", 
                              "character"))
keys_identifier = "CTUID"

SA_mode_wide <- SA_mode_wide %>%
  left_join(keys, by = keys_identifier) %>% 
  dplyr::select(-PCD)
  
respondents <- census_filtered_labeled %>% 
  mutate(PRCDDA = as.character(PRCDDA)) %>%
  left_join(SA_mode_wide, by = c("PRCDDA" = "DAUID"))
```

```{r}
respondents_filtered <- respondents %>%
  filter(
    Mob5 == 6, # Place of residence 5 years ago, code 6: Same address (dwelling)
    #MRNFY %in% c(-5,5,7,8,9), 
    !is.na(SA_pp_Transit), # Cases without accessibility values
    !is.na(SA_pp_Active), 
    !is.na(SA_pp_Car_motor),
    #!LF71 %in% c(-3,1,3,4), 
    !is.na(employed_dummy)
    )
```

Visualizing the unweighted summary: 

```{r}
respondents_filtered %>%
  dplyr::select(LOKWK_label, SA_pp_Active, SA_pp_Transit, SA_pp_Car_motor, 
                AGE, Education, Gender, Disability, Temporary_or_refugee, 
                Visible_minority, Single_parent, Children0to5, CFCNT_PP, 
                Weeks, unemployment_rate, car_share, Pop_centre, Pop_density) %>%
  summary()
```


# Models

We now develop regression models to examine the relationship between employment outcomes and neighbourhood accessibility. First, we model the link between accessibility and employment status. Next, we restrict the sample to unemployed individuals (those in the labour force but not employed) and analyse the relationship between job search engagement and neighbourhood accessibility. 

| Model | Population                       | Dependent variable                                                            |
| ----- | -------------------------------- | ----------------------------------------------------------------------------- |
| **1** | Non-disadvantaged                | Employment status (“Employed full-time” as the base category)                 |
| **2** | Disadvantaged                    | Employment status (“Employed full-time” as the base category)                 |
| **3** | Unemployed and non-disadvantaged | Job search engagement (“Yes, looked for full-time work” as the base category) |
| **4** | Unemployed and disadvantaged     | Job search engagement (“Yes, looked for full-time work” as the base category) |

Based on the literature, we include the following socioeconomic and demographic control variables in all models:

- Respondent's age and age squared (divided by 100)
- Educational background
- Number of children under age 5 in the household
- Gender
- Disability indicator
- Racialized identity
- Number of Weeks worked in the previous year
- Single-parent household
- Family size
- Neighbourhood unemployment rate (to capture job competition)
- Car share in the neighbourhood
- Population centre category
- Population density

In addition, we include the instrumental variables generated using the spatial filters. 

Because the dependent variables are categorical, and to address potential endogeneity, we apply the Two-Stage Residual Inclusion (2SRI) method. This approach involves two steps:
- First stage: Regress each endogenous accessibility variable on its instrumental variable(s) and all exogenous covariates, then extract the residuals.
- Second stage: Include these residuals as additional regressors in the main models to correct for endogeneity.

## First stage

We will start the first stage of regression by creating a formula to regress accessibility based on spatial filters and other variables. 

```{r}
design_respondents <- svydesign(ids = ~1, 
                                data = respondents_filtered, 
                                weights = ~CompW1)
```

## Active IV

Creating the first stage for cycling accessibility: 

```{r}
stage1_active <- svyglm(SA_pp_Active ~ 
                              Sf_dist_SA_pp_Active + 
                              SA_pp_Transit + 
                              `SA_pp_Car_motor` +
                              AGE + 
                              Education + 
                              Gender + 
                              Disability +
                              Temporary_or_refugee + 
                              Visible_minority + 
                              Single_parent + 
                              Children0to5 + 
                              CFCNT_PP + 
                              Weeks + 
                               
                               # Neighbourhood 
                              unemployment_rate + 
                              car_share + 
                              Pop_centre + 
                              Pop_density, design = design_respondents)

summary(stage1_active, diagnostics = TRUE)
```

```{r}
regTermTest(stage1_active, ~ Sf_dist_SA_pp_Active)
```

## Transit IV 

Creating the first stage for transit accessibility: 

```{r}
stage1_transit <- svyglm(SA_pp_Transit ~ 
                              SA_pp_Active + 
                              Sf_dist_SA_pp_Transit + 
                              `SA_pp_Car_motor` +
                              AGE + 
                              Education + 
                              Gender + 
                              Disability +
                              Temporary_or_refugee + 
                              Visible_minority + 
                              Single_parent + 
                              Children0to5 + 
                              CFCNT_PP + 
                              Weeks + 
                               
                              # Neighbourhood 
                              unemployment_rate + 
                              car_share + 
                              Pop_centre + 
                              Pop_density, design = design_respondents)

summary(stage1_transit, diagnostics = TRUE)
```

```{r}
regTermTest(stage1_transit, ~ Sf_dist_SA_pp_Transit)
```

## Car IV 

Creating the first stage for car/motorized accessibility: 

```{r}
stage1_car <- svyglm(SA_pp_Car_motor ~ 
                              SA_pp_Active + 
                              SA_pp_Transit + 
                              Sf_dist_SA_pp_Car_motor +
                              AGE + 
                              Education + 
                              Gender + 
                              Disability +
                              Temporary_or_refugee + 
                              Visible_minority + 
                              Single_parent + 
                              Children0to5 + 
                              CFCNT_PP + 
                              Weeks + 
                               
                               # Neighbourhood 
                              unemployment_rate + 
                              car_share + 
                              Pop_centre + 
                              Pop_density, design = design_respondents)

summary(stage1_car, diagnostics = TRUE)
```

```{r}
regTermTest(stage1_car, ~ Sf_dist_SA_pp_Car_motor)
```

Adding the residuals to the data frame: 

```{r}
respondents_filtered <- respondents_filtered %>%
  mutate(SA_pp_Transit_IV = predict(stage1_transit, newdata = ., type = "response"),
    SA_pp_Car_IV = predict(stage1_car, newdata = ., type = "response"),
    SA_pp_Active_IV = predict(stage1_active, newdata = ., type = "response"),
    SA_pp_Transit_Res_IV = SA_pp_Transit - SA_pp_Transit_IV,  
    SA_pp_Car_Res_IV = `SA_pp_Car_motor` - SA_pp_Car_IV,  
    SA_pp_Active_Res_IV = SA_pp_Active - SA_pp_Active_IV)
```

# Employment status

```{r}
design_employment <- svydesign(ids = ~1, data = respondents_filtered, weights = ~CompW1)

employment_formula <- as.formula(employed_dummy ~ 
                                     SA_pp_Active + 
                                     SA_pp_Transit + 
                                     SA_pp_Car_motor + 
                                     AGE + 
                                     Education + 
                                     Gender + 
                                     Disability + 
                                     Temporary_or_refugee + 
                                     Visible_minority + 
                                     Single_parent + 
                                     Children0to5 + 
                                     CFCNT_PP + 
                                     Weeks + 
                                     unemployment_rate + 
                                     car_share + 
                                     Pop_centre + 
                                     Pop_density + 
                                     SA_pp_Active_Res_IV + 
                                     SA_pp_Transit_Res_IV + 
                                     SA_pp_Car_Res_IV)
```


```{r}
employment_model <- survey::svyglm(employment_formula, 
                                   design = design_employment,
                                   family = binomial())

summary(employment_model)
```

Plotting some diagnostic graphs. The first is Cook's distance (Hint: it needs to be lower than 2): 

```{r}
#plot(svydiags::svyCooksD(employment_model))
```

Viewing the standard deviation of the residuals (Hint: no greater than 3):
 
```{r}
svydiags::svystdres(employment_model, doplot = TRUE)
```

```{r}
null_model <- svyglm(employed_dummy ~ 1, design = design_employment, family = binomial())

pseudo_r2 <- 1 - as.numeric(logLik(employment_model)) / as.numeric(logLik(null_model))

pseudo_r2
```

For each outcome, we fit the model, then iteratively remove outliers with absolute studentised residuals > 3 (using `svystdres`) to improve robustness. This is repeated up to 10 times.


```{r}
formula_current <- employment_formula
model_current <- employment_model
design_current <- design_employment
```


```{r}
max_iter <- 10
iter <- 0
outliers_removed <- 0

while (iter < max_iter) {
  # survey-weighted standardized residuals
  stdres_obj <- svystdres(model_current)
  stdres_vals <- stdres_obj$stdresids

  influential <- which(abs(stdres_vals) > 3)
  
  if (length(influential) == 0) {
    cat("No more outliers with |stdres| > 3. Stopping.\n")
    break
  }
  
  cat("Iteration", iter + 1, ": removing", length(influential), "outliers\n")
  
  keep <- !(seq_len(nrow(design_current$variables)) %in% influential)
  design_new <- subset(design_current, keep)
  
  # refit the model
  model_new <- svyglm(formula_current,
                      design = design_new)
  
  design_current <- design_new
  model_current <- model_new
  iter <- iter + 1
  outliers_removed <- outliers_removed + length(influential)
}

cat("Total outliers removed:", outliers_removed, "\n")
```

```{r}
pseudo_r2 <- 1 - as.numeric(logLik(model_current)) / as.numeric(logLik(null_model))

pseudo_r2
```

```{r}
summary(model_current)
```

Assessing the existence of multicollinearity (Tip: VIF < 10):
 
```{r}
car::vif(model_current)
```

```{r}
regTermTest(model_current, ~  SA_pp_Active + SA_pp_Transit + 
    SA_pp_Car_motor)
``` 

```{r}
svydiags::svystdres(model_current, doplot = TRUE)
```

```{r}
summary(svydiags::svystdres(model_current)$stdresids)
```

```{r}
employment_model_clean <- model_current
employment_model_clean_design <- design_current
```


```{r}
compute_or_table <- function(svy_model, 
                                      remove_intercept = TRUE, 
                                      round_digits = 3,
                                      alpha = 0.05,
                                      include_se = TRUE) {
  # Get model summary
  model_summary <- summary(svy_model)
  
  # Extract coefficients and standard errors
  coefs <- coef(svy_model)
  se <- sqrt(diag(vcov(svy_model)))
  
  # Degrees of freedom for t-distribution
  df_resid <- degf(svy_model$survey.design)
  
  # Calculate t-statistics and p-values
  t_stats <- coefs / se
  p_values <- 2 * pt(-abs(t_stats), df = df_resid)  # Two-tailed
  
  # Calculate confidence intervals using t-distribution
  t_critical <- qt(1 - alpha/2, df = df_resid)
  ci_lower <- coefs - t_critical * se
  ci_upper <- coefs + t_critical * se
  
  # Create results table
  results <- data.frame(
    Variable = names(coefs),
    Coef = coefs,
    SE = se,
    t_value = t_stats,
    OR = exp(coefs),
    OR_lower = exp(ci_lower),
    OR_upper = exp(ci_upper),
    p_value = p_values,
    stringsAsFactors = FALSE
  )
  
  # Calculate percentage change
  results$Change <- (results$OR - 1) * 100
  
  # Determine effect direction with more precise logic
  results$Effect <- with(results, {
    case_when(
      p_value >= alpha ~ "Insignificant",
      OR > 1.001 ~ "Positive",  # Small buffer for rounding
      OR < 0.999 ~ "Negative",
      TRUE ~ "No Effect"  # OR ~= 1
    )
  })
  
  # Add significance stars if desired
  results$Sig <- with(results, {
    case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE ~ ""
    )
  })
  
  # Remove intercept if requested
  if (remove_intercept && "(Intercept)" %in% results$Variable) {
    results <- results[results$Variable != "(Intercept)", ]
  }
  
  # Round numeric columns
  num_cols <- c("Coef", "SE", "t_value", "OR", "OR_lower", 
                "OR_upper", "Change")
  results[num_cols] <- lapply(results[num_cols], round, round_digits)
  results$p_value <- signif(results$p_value, 3)  # Better for p-values
  
  # Reorder columns
  col_order <- c("Variable", "Coef", if(include_se) "SE", "t_value", 
                 "OR", "OR_lower", "OR_upper", "Change", "p_value", 
                 "Sig", "Effect")
  col_order <- col_order[col_order %in% names(results)]
  
  results <- results[, col_order]
  
  # Add interpretation column
  results$Interpretation <- with(results, {
    case_when(
      Effect == "Positive" ~ sprintf(
        "%s%% increase in odds (95%% CI: %s to %s)",
        round(Change, 1), round(OR_lower, 2), round(OR_upper, 2)
      ),
      Effect == "Negative" ~ sprintf(
        "%s%% decrease in odds (95%% CI: %s to %s)",
        round(-Change, 1), round(OR_lower, 2), round(OR_upper, 2)
      ),
      TRUE ~ "No significant effect"
    )
  })
  
  return(results)
}

compute_elasticities <- function(model, alpha = 0.05) {
  mf <- model.frame(model$survey.design)
  w  <- weights(model$survey.design)
  
  pred_probs <- predict(model, type = "response")
  
  coefs <- coef(model)
  vcov_mat <- vcov(model)
  
  cont_vars <- intersect(names(coefs), colnames(mf))
  cont_vars <- cont_vars[cont_vars != "(Intercept)"]
  cont_vars <- cont_vars[sapply(mf[cont_vars], function(x) length(unique(na.omit(x))) > 2)]
  cont_vars <- cont_vars[cont_vars %in% names(coefs)]
  
  elas <- sapply(cont_vars, function(v) {
    x_mean <- weighted.mean(mf[[v]], w, na.rm = TRUE)
    p_mean <- weighted.mean(pred_probs, w, na.rm = TRUE)
    elas_val <- coefs[v] * x_mean / p_mean
    elas_se <- sqrt((x_mean / p_mean)^2 * vcov_mat[v, v])
    CI_lower <- elas_val - qnorm(1 - alpha/2) * elas_se
    CI_upper <- elas_val + qnorm(1 - alpha/2) * elas_se
    return(c(Elasticity = elas_val, `2.5%` = CI_lower, `97.5%` = CI_upper))
  })
  
  elas_df <- as.data.frame(t(elas))
  colnames(elas_df) <- c("Elasticity", "2.5%", "97.5%")
  elas_df$Variable <- rownames(elas_df)
  
  elas_df$Effect <- ifelse(elas_df$`2.5%` > 0, "Positive",
                           ifelse(elas_df$`97.5%` < 0, "Negative", "Insignificant"))
  
  elas_df <- elas_df[, c("Variable", "Elasticity", "2.5%", "97.5%", "Effect")]
  
  numeric_cols <- c("Elasticity", "2.5%", "97.5%")
  elas_df[numeric_cols] <- round(elas_df[numeric_cols], 3)
  rownames(elas_df) <- NULL
  
  return(elas_df)
}

compute_marginal_effects <- function(model, categorical_vars, alpha = 0.05) {
  # Extract survey design and weights
  design <- model$survey.design
  mf <- model.frame(design)           # original data (with factors)
  w  <- weights(design)               # survey weights
  
  # Ensure categorical_vars are factors in the model frame
  categorical_vars <- intersect(categorical_vars, names(mf))
  if (length(categorical_vars) == 0) 
    stop("None of the supplied variables are factors in the model frame.")
  
  # Model matrix and coefficient vectors
  X_orig <- model.matrix(model)       # includes intercept and all dummy columns
  beta   <- coef(model)
  vcov_mat <- vcov(model)             # full variance‑covariance matrix
  
  # For each categorical variable, identify its dummy columns in X_orig
  colnames_X <- colnames(X_orig)
  
  results <- list()   # store results for each level
  
  for (var in categorical_vars) {
    # Get levels and contrast coding from the factor
    fac <- mf[[var]]
    if (!is.factor(fac)) stop(paste(var, "is not a factor. Convert it before fitting."))
    levs <- levels(fac)
    ref <- levs[1]                     # assumes treatment contrasts (R default)
    non_ref <- levs[-1]
    
    # Find dummy column indices for this factor (format "varNameLevel")
    pattern <- paste0("^", var)
    dummy_cols <- grep(pattern, colnames_X, value = TRUE)
    # Map each level to its column name (if it exists)
    level_to_col <- setNames(dummy_cols, gsub(pattern, "", dummy_cols))
    # Only keep levels that actually appear in the model matrix
    available_levels <- intersect(non_ref, names(level_to_col))
    if (length(available_levels) == 0) {
      warning(paste("No dummy columns found for", var, "- skipping."))
      next
    }
    
    for (lev in available_levels) {
      col_lev <- level_to_col[[lev]]
      
      # --- 1. Construct X_L and X_R matrices (same as X_orig but with dummy columns adjusted)
      X_L <- X_R <- X_orig
      # For the factor's dummy columns: set all to 0 for reference
      X_R[, dummy_cols] <- 0
      # For X_L: set the dummy of the target level to 1, others to 0
      X_L[, dummy_cols] <- 0
      X_L[, col_lev] <- 1
      
      # --- 2. Predicted probabilities under each scenario
      eta_L <- X_L %*% beta
      eta_R <- X_R %*% beta
      pred_L <- plogis(eta_L)
      pred_R <- plogis(eta_R)
      
      # Weighted average difference (marginal effect)
      me <- weighted.mean(pred_L - pred_R, w, na.rm = TRUE)
      
      # --- 3. Gradient vector for delta method
      # derivative for each coefficient k:
      # d_i_k = pred_L*(1-pred_L)*X_L[i,k] - pred_R*(1-pred_R)*X_R[i,k]
      # then grad[k] = weighted.mean( d_i_k )
      grad <- numeric(length(beta))
      for (k in seq_along(beta)) {
        d <- pred_L * (1 - pred_L) * X_L[, k] - pred_R * (1 - pred_R) * X_R[, k]
        grad[k] <- weighted.mean(d, w, na.rm = TRUE)
      }
      names(grad) <- names(beta)
      
      # Standard error using delta method
      se <- sqrt(t(grad) %*% vcov_mat %*% grad)
      
      # Confidence interval
      z <- qnorm(1 - alpha/2)
      ci_lower <- me - z * se
      ci_upper <- me + z * se
      
      # Interpretation
      effect <- ifelse(ci_lower > 0, "Positive",
                       ifelse(ci_upper < 0, "Negative", "Insignificant"))
      
      # Store results
      results[[length(results)+1]] <- data.frame(
        Variable = paste0(var, ": ", lev, " vs ", ref),
        MarginalEffect = me,
        Lower = ci_lower,
        Upper = ci_upper,
        Effect = effect,
        stringsAsFactors = FALSE
      )
    }
  }
  
  if (length(results) == 0) stop("No marginal effects could be computed.")
  
  final <- do.call(rbind, results)
  # Round numeric columns
  final$MarginalEffect <- round(final$MarginalEffect, 3)
  final$Lower <- round(final$Lower, 3)
  final$Upper <- round(final$Upper, 3)
  
  # Rename columns for clarity
  names(final) <- c("Variable", "Marginal Effect", "2.5%", "97.5%", "Effect")
  return(final)
}
```

Calculation of the odds ratio ($e^β$): 
 
```{r}
OR_table_employment <- compute_or_table(employment_model_clean)
OR_table_employment
```

Visualizing elasticity (the impact of a 1% increase in the outcome for numerical variables):

```{r}
elasticity_table_employment <- compute_elasticities(employment_model_clean)
elasticity_table_employment
```

Viewing the marginal effect (for categorical variables): 

```{r}
categorical_vars  <- c("Education", "Gender",
                                     "Disability",
                                     "Temporary_or_refugee",
                                     "Visible_minority",
                                     "Single_parent",
                                     "Pop_centre")

margins_employment <- compute_marginal_effects(employment_model_clean, 
                                               categorical_vars)
margins_employment
```

# Job seeking

For the second outcome, we need to filter the respondents to left only unemployed people:

```{r}
respondents_filtered_lokwk <- respondents_filtered %>%
  filter(LOKWK_label %in% c("Yes (for full-time work)", 
                            "Yes (for part-time work)", 
                            "No"),
         employed_dummy == 0) %>% 
  mutate(LOKWK_label = factor(LOKWK_label, levels = c("No",
                                                      "Yes (for part-time work)", 
                                                      "Yes (for full-time work)")))
```

```{r}
summary(respondents_filtered_lokwk$LOKWK_label)
```

Creating the formula: 

```{r}
job_seeking_formula <- as.formula(LOKWK_label ~ 
                                     SA_pp_Active + 
                                     SA_pp_Transit + 
                                     SA_pp_Car_motor + 
                                     AGE + 
                                     Education + 
                                     Gender + 
                                     Disability + 
                                     Temporary_or_refugee + 
                                     Visible_minority + 
                                     Single_parent + 
                                     Children0to5 + 
                                     CFCNT_PP + 
                                     Weeks + 
                                     unemployment_rate + 
                                     car_share + 
                                     Pop_centre + 
                                     Pop_density + 
                                     SA_pp_Active_Res_IV + 
                                     SA_pp_Transit_Res_IV + 
                                     SA_pp_Car_Res_IV)
```

```{r}
jseek_model <- vglm(job_seeking_formula,
                          family = multinomial(refLevel = 1),
                          data = respondents_filtered_lokwk,
                          weights = CompW1)

summary(jseek_model)
```

Assessing the existence of multicollinearity (Tip: VIF < 10):

```{r}
# m <- lm(job_seeking_formula, 
#                data = respondents_filtered_lokwk, 
#                weights = respondents_filtered_lokwk$CompW1)
# 
# car::vif(jseek_model)
```

Comparing our model with a null model:

```{r}
jseek_ndis_model_nulo <- vglm(update(job_seeking_formula, LOKWK_label ~ 1),
                          family = multinomial(),
                          data = respondents_filtered_lokwk,
                          weights = CompW1)
```

ANOVA: 

```{r}
anova(jseek_model, jseek_ndis_model_nulo, type = 'I')
```

```{r}
car::Anova(jseek_model)
```

Calculation of the odds ratio ($e^β$): 

```{r}
make_multinom_OR_table <- function(model) {
    coef_mat <- coef(summary(model))
  
  est <- coef_mat[, "Estimate"]
  se  <- coef_mat[, "Std. Error"]
  p   <- coef_mat[, "Pr(>|z|)"]
  
  OR <- exp(est)
  CI_low  <- exp(est - 1.96 * se)
  CI_high <- exp(est + 1.96 * se)
  
  df <- tibble(raw_term = rownames(coef_mat),
    OR = OR,
    CI_low = CI_low,
    CI_high = CI_high,
    p_value = p)
  
  categories <- model@misc$ynames       
  ref_cat <- categories[1]             
  
  df <- df %>%
    mutate(Variable = str_replace(raw_term, ":(.*)$", ""),
      Cat = str_extract(raw_term, "(?<=:)\\d+"),
      Cat = as.integer(Cat),
      Comparison = paste0(categories[Cat + 1], " vs ", ref_cat)) %>%
    dplyr::select(Variable, Comparison, OR, p_value, CI_low, CI_high)
  
  df <- df %>%
    rowwise() %>%
    mutate(Interpretation = case_when(
        is.na(p_value) ~ "",
        p_value > 0.05 ~ "No significant effect",
        OR > 1 ~ "Increase in odds",
        OR < 1 ~ "Decrease in odds",
        TRUE ~ "No effect")) %>%
    ungroup()
  
  return(df)
}
```

```{r}
OR_table_jseek_model <- make_multinom_OR_table(jseek_model)
OR_table_jseek_model
```

# Descriptive analysis

Before we perform the second stage, we will visualize some descriptive statistics about the variables used in the model.

```{r}
# Long format table
total_pop_rounded <- respondents_filtered %>%
  summarise(total = sum(CompW1, na.rm = TRUE)) %>%
  pull(total) %>%
  { round(. / 5) * 5 }

total_pop_study_area <- census_vetting %>%
  summarise(total = sum(CompW1, na.rm = TRUE)) %>%
  pull(total)

num_vars <- c("SA_pp_Active", "SA_pp_Transit", "SA_pp_Car_motor",
  "AGE", "Children0to5", "CFCNT_PP", "Weeks",
  "unemployment_rate", "car_share", "Pop_density")

cat_vars <- c("Education", "Gender", "Disability", "Temporary_or_refugee",
              "Visible_minority", "Single_parent", "Pop_centre", "LOKWK_label", "LBR_FORC_label")

num_table <- respondents_filtered %>%
  dplyr::select(all_of(num_vars), CompW1) %>%
  pivot_longer(cols = all_of(num_vars),
    names_to = "Variable",
    values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(Unweighted_n = sum(!is.na(Value)),
            Weighted_pop = sum(CompW1, na.rm = TRUE),
            Total_pop_Study_area = total_pop_study_area, 
    Weighted_mean = sum(Value * CompW1, na.rm = TRUE) / sum(CompW1, na.rm = TRUE),
    Weighted_sd = sqrt(sum(CompW1 * (Value - Weighted_mean)^2, na.rm = TRUE) / sum(CompW1, na.rm = TRUE)),
    Weighted_nonzero = sum((Value > 0) * CompW1, na.rm = TRUE),
    .groups = "drop") %>%
  mutate(across(c(Weighted_mean, Weighted_sd, Weighted_nonzero), ~ round(., 3)))

cat_table <- respondents_filtered %>%
  dplyr::select(all_of(cat_vars), CompW1) %>%
  pivot_longer(cols = all_of(cat_vars),
    names_to = "Variable",
    values_to = "Level") %>%
  filter(!is.na(Level)) %>%
  group_by(Variable, Level) %>%
  summarise(Unweighted_n = n(),
    Weighted_sum = sum(CompW1, na.rm = TRUE),
    .groups = "drop") %>%
  mutate(Weighted_rounded = round(Weighted_sum / 5) * 5,
    Proportion = Weighted_rounded / total_pop_rounded) %>%
  mutate(Proportion = round(Proportion, 4), 
         Total_pop_Study_area = total_pop_study_area)

cat_table <- cat_table %>%
  mutate(Description = case_when(Variable == "Education" ~ "Highest level of education",
                                 Variable == "Gender" ~ "Gender identity",
                                 Variable == "Disability" ~ "Disability status",
                                 Variable == "Temporary_or_refugee" ~ "Temporary or refugee status",
                                 Variable == "Visible_minority" ~ "Visible minority status",
                                 Variable == "Single_parent"  ~ "Single parent status",
                                 Variable == "Pop_centre" ~ "Population centre type",
                                 Variable == "LOKWK_label" ~ "Looked for work",
                                 Variable == "LBR_FORC_label" ~ "Labour fource",
                                 TRUE ~ Variable))

cat_table <- cat_table %>%
  dplyr::select(Variable, Description, Level, Unweighted_n, Weighted_sum, 
         Weighted_rounded, Proportion, everything())
```

```{r}
get_p_percentage <- function(nonzero_pop) {
  case_when(nonzero_pop < 10 ~ 0.1667, # 16.67%
    nonzero_pop < 20 ~ 0.1667, # 16.67%
    nonzero_pop < 40 ~ 0.10, # 10%
    nonzero_pop < 80 ~ 0.05, # 5%
    nonzero_pop < 160 ~ 0.025, # 2.5%
    nonzero_pop < 320 ~ 0.0125, # 1.25%
    nonzero_pop < 640 ~ 0.00625, # 0.625%
    nonzero_pop < 1280 ~ 0.00313, # 0.313%
    TRUE ~ 0.00156 # 0.156%
  )
}

round_statistic <- function(value, nonzero_pop) {
  if (is.na(value) || value == 0) return(value)
  
  p_percent <- get_p_percentage(nonzero_pop)
  allowed_error <- abs(value) * p_percent
  rounding_base <- 2 * allowed_error
  
  if (rounding_base == 0) return(value)
  
  return(round(value / rounding_base) * rounding_base)
}
```


```{r}
num_table_rounded <- num_table %>%
  group_by(Variable) %>%
  mutate(Weighted_mean_rounded = mapply(round_statistic, Weighted_mean, Weighted_nonzero),
    Weighted_sd_rounded   = mapply(round_statistic, Weighted_sd,   Weighted_nonzero)) %>%
  ungroup()

num_table_rounded <- num_table_rounded %>% 
  mutate(Description = case_when(Variable == "AGE" ~ "Age",
                   Variable == "CFCNT_PP" ~ "Family size",
                   Variable == "Children0to5" ~ "Number of children aged 0 to 5",
                   Variable == "Pop_density" ~ "Population Density of the Dissemination Area",
                   Variable == "SA_pp_Car_motor" ~ "Car accessibility",
                   Variable == "SA_pp_Transit" ~ "Transit accessibility",
                   Variable == "SA_pp_Active" ~ "Active accessibility",
                   Variable == "Weeks" ~ "Number of Weeks worked in the previous year",
                   Variable == "car_share" ~ "Percentual de car commuters in the DA",
                   Variable == "unemployment_rate" ~ "Unemployment rate of the DA", 
                   TRUE ~ Variable))


num_table_rounded <- num_table_rounded %>% 
   relocate(Description, .after = 1) 
```


# Confidentiality vetting

Confidentiality vetting is the process of reviewing the results to be
released by the Research Data Center to ensure that confidentiality
risks for StatCan respondents are minimized.

The following rules apply to our results:

-   *Statistics must not be released for identifiable areas with less
    than 40 persons (*$\sum{CompW1} ≥ 40$). Statistics must not be
    released for identifiable areas with less than 40 persons. This
    condition is usually met with geography levels used at the RDCs. The
    population threshold can be applied to the (long form) weighted
    population estimate.
-   *All statistics must be based on weighted data.*
-   *Statistics must be based on at least 5 respondents, whose sum of
    weights is at least 10*.
-   *Statistics (Means and total sums) are subject to a simplistic
    percent rounding by deterministically rounding to the nearest
    multiple of a rounding base.* The rounding base is set so that the
    rounded value is never more than p% away from the unrounded value.
    The value of p% is determined by the associated nonzero population
    size of the mean or total. It is the weighted number of records with
    a nonzero value. The table below outlines the maximum percentage
    difference (p%) from the unrounded value, based on its associated
    nonzero population size.

| Nonzero Pop. size | \<10 | 10-\<20 | 20-\<40 | 40-\<80 | 80-\<160 | 160-\<320 | 320-\<640 | 640-\<1280 | \>=1280 |
|----|----|----|----|----|----|----|----|----|----|
| p% | 16.67% | 16.67% | 10% | 5% | 2.5% | 1.25% | 0.625% | 0.313% | 0.156% |


# Export files 

Descriptive measures:

```{r export-original-files}
# Descriptive analysis (numerical variables)
write.csv(num_table_rounded, paste0(diretorio_export, "descriptive_analysis_numeric_original.csv"), row.names=FALSE)

da_support_excel <- createWorkbook()

addWorksheet(da_support_excel, "Sheet1")

writeData(da_support_excel, "Sheet1", x = num_table_rounded)  

writeComment(da_support_excel, "Sheet1", col = 1, row = 1,  comment = createComment("Variable name used in the model.", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 2, row = 1,  comment = createComment("Variable label.", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 3, row = 1,  comment = createComment("Unweighted population", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 4, row = 1,  comment = createComment("Weighted Population (sum of CompW1)", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 5, row = 1,  comment = createComment("Total GTAH population", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 6, row = 1,  comment = createComment("Mean value of the variable", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 7, row = 1,  comment = createComment("Standard deviation of the variable", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 8, row = 1,  comment = createComment("Non-zero population size of the variable", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 9, row = 1,  comment = createComment("Rounded (based on the non-zero population size) mean of the variable", author = "Author"))
writeComment(da_support_excel, "Sheet1", col = 10, row = 1,  comment = createComment("Rounded (based on the non-zero population size) standard deviation of the variable", author = "Author"))

cols_with_values_general <- which(sapply(num_table_rounded, 
                                         function(x) any(!is.na(x))))

addStyle(da_support_excel, sheet = c("Sheet1"),
         style = createStyle(fgFill = "yellow"),
         rows = 1:(nrow(num_table_rounded)+1),
         cols = c(1,2,9,10),
         gridExpand = TRUE)

saveWorkbook(da_support_excel, paste0(diretorio_export, "descriptive_analysis_numeric_support.xlsx"), overwrite = TRUE)

num_table_rounded_release <- num_table_rounded %>% 
  dplyr::select(Variable, Description, Weighted_mean_rounded, Weighted_sd_rounded)

da_release_excel <- createWorkbook()

addWorksheet(da_release_excel, "Sheet1")

writeData(da_release_excel, "Sheet1", 
          x = num_table_rounded_release)  

writeComment(da_release_excel, "Sheet1", col = 1, row = 1,  comment = createComment("Variable name used in the model.", author = "Author"))
writeComment(da_release_excel, "Sheet1", col = 2, row = 1,  comment = createComment("Variable label.", author = "Author"))
writeComment(da_release_excel, "Sheet1", col = 3, row = 1,  comment = createComment("Rounded (based on the non-zero population size) mean of the variable", author = "Author"))
writeComment(da_release_excel, "Sheet1", col = 4, row = 1,  comment = createComment("Rounded (based on the non-zero population size) standard deviation of the variable", author = "Author"))

cols_with_values_general <- which(sapply(num_table_rounded_release, 
                                         function(x) any(!is.na(x))))
addStyle(da_release_excel, sheet = c("Sheet1"),
         style = createStyle(fgFill = "yellow"),
         rows = 1:(nrow(num_table_rounded_release)+1),
         cols = c(1:4),
         gridExpand = TRUE)

saveWorkbook(da_release_excel, paste0(diretorio_export, "descriptive_analysis_numeric_release.xlsx"), overwrite = TRUE)
```


```{r}
# Descriptive analysis (categorical variables)
write.csv(cat_table, paste0(diretorio_export, "descriptive_analysis_categorical_original.csv"), 
          row.names = FALSE)

da_cat_support <- createWorkbook()
addWorksheet(da_cat_support, "Sheet1")
writeData(da_cat_support, "Sheet1", x = cat_table)

writeComment(da_cat_support, "Sheet1", col = 1, row = 1, 
             comment = createComment("Variable name used in the model.", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 2, row = 1, 
             comment = createComment("Descriptive label of the variable.", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 3, row = 1, 
             comment = createComment("Category level.", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 4, row = 1, 
             comment = createComment("Unweighted count.", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 5, row = 1, 
             comment = createComment("Weighted sum (CompW1).", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 6, row = 1, 
             comment = createComment("Weighted sum rounded to nearest 5.", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 7, row = 1, 
             comment = createComment("Proportion (rounded sum / total rounded population).", author = "Author"))
writeComment(da_cat_support, "Sheet1", col = 8, row = 1, 
             comment = createComment("Total GTAH population.", author = "Author"))

addStyle(da_cat_support, sheet = "Sheet1",
         style = createStyle(fgFill = "yellow"),
         rows = 1:(nrow(cat_table) + 1),
         cols = c(1, 2, 3, 7),
         gridExpand = TRUE)

saveWorkbook(da_cat_support, 
             paste0(diretorio_export, "descriptive_analysis_categorical_support.xlsx"), 
             overwrite = TRUE)

cat_table_release <- cat_table %>%
  dplyr::select(Variable, Description, Level, Proportion)

da_cat_release <- createWorkbook()
addWorksheet(da_cat_release, "Sheet1")
writeData(da_cat_release, "Sheet1", x = cat_table_release)

writeComment(da_cat_release, "Sheet1", col = 1, row = 1, 
             comment = createComment("Variable name used in the model.", author = "Author"))
writeComment(da_cat_release, "Sheet1", col = 2, row = 1, 
             comment = createComment("Descriptive label of the variable.", author = "Author"))
writeComment(da_cat_release, "Sheet1", col = 3, row = 1, 
             comment = createComment("Category level.", author = "Author"))
writeComment(da_cat_release, "Sheet1", col = 4, row = 1, 
             comment = createComment("Proportion of total population.", author = "Author"))

addStyle(da_cat_release, sheet = "Sheet1",
         style = createStyle(fgFill = "yellow"),
         rows = 1:(nrow(cat_table_release) + 1),
         cols = 1:4,
         gridExpand = TRUE)

saveWorkbook(da_cat_release, 
             paste0(diretorio_export, "descriptive_analysis_categorical_release.xlsx"), 
             overwrite = TRUE)
```

Exporting models: 

```{r}
# First stage regressions
capture.output( summary(stage1_active),
  file = paste0(diretorio_export, "stage1_active.txt"))

capture.output( summary(stage1_transit),
  file = paste0(diretorio_export, "stage1_transit.txt"))

capture.output( summary(stage1_car),
  file = paste0(diretorio_export, "stage1_car.txt"))

# Second stage regressions
capture.output(summary(employment_model_clean),
  file = paste0(diretorio_export, "model_employment_status.txt"))

capture.output(summary(jseek_model),
  file = paste0(diretorio_export, "model_job_seeking.txt"))
```

For future consultations: 

```{r}
saveRDS(employment_model_clean, paste0(diretorio_export,"employment_model_clean.rds"))
saveRDS(employment_model_clean_design, paste0(diretorio_export,"employment_model_clean_design.rds"))

saveRDS(jseek_model, paste0(diretorio_export,"jseek_model.rds"))
saveRDS(respondents_filtered, paste0(diretorio_export,"respondents_filtered_employment.rds"))
saveRDS(respondents_filtered_lokwk, paste0(diretorio_export,"respondents_filtered_lokwk.rds"))
```

```{r}
write.csv(margins_employment, paste0(diretorio_export, "margins_employment.csv"), row.names = FALSE)

write.csv(elasticity_table_employment, paste0(diretorio_export, "elasticity_table_employment.csv"), row.names = FALSE)

write.csv(OR_table_employment, paste0(diretorio_export, "OR_table_employment.csv"), row.names = FALSE)

write.csv(OR_table_jseek_model, paste0(diretorio_export, "OR_table_jseek_model.csv"), row.names = FALSE)
```

