---
title: "Calculating the accessibility and spatial availability of jobs (_Dissemination area_ level)"
runningheader: "a methodology for Canadian regions based on the 2021 Census of Population" # only for pdf output
subtitle: "a methodology for Canadian regions based on the 2021 Census of Population" # only for html output
author: "Bruno Santos & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r clean-workspace, include=FALSE}
# cleaning objects from the workspace 
rm(list = ls())
```

```{r setup, include=FALSE}
# layout configuration 
library(tufte)
library(knitr)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

This Rmarkdown file is part of the
[**CommuteCA**](https://github.com/dias-bruno/CommuteCA) package.
This package was created in conjunction with the office of the
[*Research Data Center* at *McMaster
University*](https://rdc.mcmaster.ca/), the [*Sherman Centre for Digital
Scholarship*](https://scds.ca/) and the [*Mobilizing
Justice*](https://mobilizingjustice.ca/)[^1].

[^1]: The Mobilizing Justice project is a multidisciplinary and
    multi-sector collaboration with the objective of understand and
    address transportation poverty in Canada and to improve the
    well-being of Canadians at risk of transport poverty. The Social
    Sciences and Humanities Research Council (SSRHC) has provided
    funding for the project, which was created by an unprecedented
    alliance of academics from various Canadian provinces and
    institutions, transportation firms, and nonprofit organizations

The [**CommuteCA**](https://github.com/dias-bruno/CommuteCA) R
package was created to develop standardized methods for transport
analysis in research, particularly for analysis using the [*2021 Census of
Population*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm) from Statistics Canada. We focused our efforts on the [_Commuting Reference Guide_](https://www12.statcan.gc.ca/census-recensement/2021/ref/98-500/011/98-500-x2021011-eng.cfm),
which provides valuable variables and information on commuting for the Canadian population aged 15 and older living in private households. 

Considering the number of workers and employment opportunities obtained
from the [*2021 Census of
Population*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm),
this R markdown aims to create a methodology to obtain Hansen-type
accessibility [@hansen1959] and spatial accessibility [@soukhov2023],
for all Canadian provinces and territories, considering different modes
of travel.

## Suggested Readings

-   Soukhov, Anastasia, Antonio Paez, Christopher D Higgins, and Moataz
    Mohamed. 2023."Introducing Spatial Availability, a
    Singly-Constrained Measure of Competitive Accessibility." *Plos One*
    18 (1): e0278468.
-   Soukhov, A., Páez, A. (2024). Accessibility analysis for planning
    applications (Report No. MJ-A2-0002). Mobilizing Justice.
    <https://github.com/soukhova/MJ-Accessibility-Blogs>
-   Pereira, Rafael H. M.; Herszenhut, Daniel. Introduction to urban
    accessibility: a practical guide with R. Rio de Janeiro: Ipea, 2023.
    152 p. ISBN: 978-65-5635-065-3. DOI:
    <http://dx.doi.org/9786556350653>.

## Data

The dataset used in this demonstration is test data produced to
replicate the variables available in the original Census of Population.
The test data contains 200,000 rows and 17 columns. As in the original census data,
each row refers to a respondent and each column refers to a
variable[^2].\
The creation of test data was necessary because the surveys
provided by Statistics Canada are confidential and cannot be accessed
outside of a Research Data Center.

[^2]: You can check out more information about the Census on the
    [Dictionary
    website](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/index-eng.cfm).

If you want to work with the original Census dataset, the process for
obtaining the accessibility measures will be the same as for the
test data, except that you will have to update the address of the
file in the chunk[^3] called *load-census-data*.

[^3]: A code chunk is an executable part of the R code

For this R markdown, we'll use the following variables[^4]:

[^4]: The explanation of each variable can be found in the [*2021 Census
    of Population's
    website*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm).

|              |                                                                                                                                 |
|------------------------|------------------------------------------------|
| **Variable** | **Description**                                                                                                                 |
| PRCDDA       | Refers to the dissemination area (DA) of current residence.                                                                     |
| Pr           | Refers to the province or territory of current residence.                                                                       |
| CMA          | Census metropolitan area or census agglomeration of current residence.                                                          |
| PCD          | Census division of current residence.                                                                                           |
| CompW1       | Weight for the households and dwellings universes.                                                                              |
| LBR_FORC     | This variable refers to whether a person was employed, unemployed or not in the labour force.                                   |
| PWDA         | Place of work dissemination area.                                                                                               |
| PWPR         | Place of work province.                                                                                                         |
| PWCMA        | Census metropolitan area or census agglomeration of place of work.                                                              |
| PWCD         | Place of work census division.                                                                                                  |
| PWDUR        | Commuting duration, it refers to the length of time, in minutes, usually required by a person to travel to their place of work. |
| PwMode       | Main mode of commuting' refers to the main mode of transportation a person uses to travel to their place of work.               |

: Census variables used in this R markdown.

A dissemination area (DA) is a small, relatively stable geographic unit
composed of one or more adjacent dissemination blocks. It is the
smallest standard geographic area for which all census data are
disseminated. DAs cover all Canadian territory.

We will also use the travel time table created in the previous step, in
the R markdown called *03_travel_times.Rmd*.[^5].

[^5]: Although we have provided test census data for all Canadian
    provinces and territories, we have produced a travel table only for
    the city of Toronto. So if you want to test the methodology for
    other cities or regions, you'll have to generate a travel time for
    the corresponding study area with *01_travel_times.Rmd*

The travel times table has the following variables:

| **Variable** | **Description**                                                                   |
|----------------|--------------------------------------------------------|
| PRCDDA       | Refers to the DA defined as the origin.                                           |
| CMAUID       | Census metropolitan area or census agglomeration of the DA defined as the origin. |
| PWDA         | Refers to the DA defined as the destination.                                      |
| travel_time  | Estimated travel time in minutes from origin to destination.                      |
| PwMode       | Transportation mode used to calculate the travel time.                            |

: Travel time table variables used in this R markdown.

# Spatial filter 

# Let's code!

<!-- The methodology for obtaining the accessibility measures using the Canadian -->
<!-- Census consists of the following steps: -->

<!-- 1.  First, we estimate the travel time for each transportation mode using the [{r5r}](https://cloud.r-project.org/web/packages/fitdistrplus/index.html) package [@Pereira2021], considering all possible combinations of dissemination areas of the selected study area. The methodology used to obtain the travel time table is in the *01_travel_times.Rmd*.  -->

<!-- 2.  After this, we extract the commute time to work declared by the Census respondents, in order to calibrate the impedance functions for each transportation mode. -->

<!-- 3.  We then apply the impedance functions to the travel time table. -->

<!-- 4.  Next, we create a land use table. This table contains the labour force population and the number of job opportunities for each dissemination area. -->

<!-- 5.  We join the travel time and labour force tables, creating the accessibility table. This last table has all the data needed to carry out the accessibility analysis.  -->

<!-- 6.  Finally, we perform the Hansen accessibility type and the spatial availability. -->

Load the packages:

```{r load-packages}
library(CommuteCA)
library(dplyr)# A Grammar of Data Manipulation 
library(here) # enable easy file referencing in project-oriented workflows
library(ggplot2) # Create Elegant Data Visualizations Using the Grammar of Graphics
library(RColorBrewer) # color schemes for maps (and other graphics) designed
library(sf) # support for simple features, a standardized way to encode spatial vector data
library(tmap) # thematic maps
library(tidyr) # tidying data
library(purrr) # Filter data 
```

## Reading and processing the census file

Reading census data and creating a R data frame[^1]:

[^1]: For this demonstration, we will use the case of the city of
    Toronto. If you want to use the other options, the original data
    from the Census in an RDC office or the test data for all
    locations in Canada, *update* the address in the chunk.
    
```{r}
# For the original Census data set
files_address <- paste0(here(),"/data-ignore-inputs/census_test_toronto.csv") # insert-the-RDC-address!!!
census <- read.csv(files_address, header = TRUE)
```

If the original Census dataset available in the RDC is not in .csv (comma-separated values) format but is instead provided in other formats such as SPSS, SAS, or SAS Data, you can use the foreign package (a built-in R library) to import it:

```{r}
# library(foreign)
# foreign::read.dta(files_address) # For Stata
# foreign::read.spss(files_address) # For SPSS
# There are many other options! You can search for this library in the 'Packages' window and explore additional functions for reading your file. 
```

Or if you have installed *CommuteCA*, you can access the *test* data:

```{r load-census-data}
# data("census_test_toronto")
# census <- census_test_toronto
```

| ⚠️**NOTE:**  If the code above did not run correctly, you probably are experiencing a file address error. Try to identify the correct address and update the chunk named `census-file-address` to continue.

It's possible to filter the data frame by administrative (province
and/or census division) and/or statistical (census metropolitan areas
and census agglomerations)[^10]. The chunk below shows how to make this
procedure:

[^10]: As said before, we will perform our analysis for the city of
    Toronto. If you want to select a specific area to work, uncomment
    the code above that applies to your case and select the apropriate
    code of your interest unit. Please, check the dictionary to have
    more informations about the [provinces
    code](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/tab/index-eng.cfm?ID=t1_8),
    [census
    divisions](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo008),
    and [census metropolitan
    areas](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo009)

```{r filter-boundaries}
code <- 3520 # census division of Toronto

# Filtering by census division
 census_filtered <- census %>%
                    filter(PCD == code) # Only select respondents who live in  Toronto
```

According to the census code book, the variable 'PwMode' has the
following possible values:

-   -3: Not applicable.
-   1: Car, truck or van - as a driver.
-   2: Car, truck or van - as a passenger.
-   3: Bus.
-   4: Subway or elevated rail.
-   5: Light rail, streetcar or commuter train.
-   6: Passenger ferry.
-   7: Walked.
-   8: Bicycle.
-   9: Motorcycle, scooter or moped.
-   10: Other method.

We'll rename the travel modes to facilitate the readability of the data.
Additionally, we'll remove from our analysis travel modes signed as
'Other methods':

According to the Census 2021, an 'Employed' person refers to those who,
during the reference period, had a labour force status of
'employed.'[^11]. In addition, the labour force is made up of the
employed population and the unemployed population. This variable can
assume the values:

[^11]: Those who, during the reference period: (a) Did any work at all
    at a job or business, that is, paid work in the context of an
    employer-employee relationship, or self-employment. This also
    includes persons who did unpaid family work, which is defined as
    unpaid work contributing directly to the operation of a farm,
    business or professional practice owned and operated by a related
    member of the same household; or (b) Had a job but were not at work
    due to factors such as their own illness or disability, personal or
    family responsibilities, vacation or a labour dispute. This category
    excludes persons not at work because they were on layoff or between
    casual jobs, and those who did not then have a job (even if they had
    a job to start at a future date).

-   -3: Not Applicable, \< 15 years
-   1: In Labour Force, Employed
-   2: In Labour Force, Unemployed
-   3: Not in Labour Force

```{r lf-selection}
census_filtered <- census_filtered %>% 
                   filter(LBR_FORC == 1 | LBR_FORC == 2)
```

According to the Census 2021, an 'Employed' person refers to those who,
during the reference period, had a labour force status of
'employed.'[^6]. In addition, the labour force is made up of the
employed population and the unemployed population. This variable can
assume the values:

[^6]: Those who, during the reference period: (a) Did any work at all at
    a job or business, that is, paid work in the context of an
    employer-employee relationship, or self-employment. This also
    includes persons who did unpaid family work, which is defined as
    unpaid work contributing directly to the operation of a farm,
    business or professional practice owned and operated by a related
    member of the same household; or (b) Had a job but were not at work
    due to factors such as their own illness or disability, personal or
    family responsibilities, vacation or a labour dispute. This category
    excludes persons not at work because they were on layoff or between
    casual jobs, and those who did not then have a job (even if they had
    a job to start at a future date).

-   -3: Not Applicable, \< 15 years (already previously excluded from
    our dataset)
-   1: In Labour Force, Employed
-   2: In Labour Force, Unemployed
-   3: Not in Labour Force (already previously excluded from our
    dataset)

According to the census code book, the variable 'GENDER_BINARY' has the
following possible values:[^7]

[^7]: Given that the non-binary population is small, data aggregation to
    a two-category gender variable is sometimes necessary to protect the
    confidentiality of responses provided. In these cases, individuals
    in the category “non-binary persons” are distributed into the other
    two gender categories and are denoted by the “+” symbol.

-   1: Woman+.
-   2: Man+.

According to the census code book, the variable 'GENDER_BINARY' has the
following possible values:

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Concept not applicable (territories or reserves) - not in low
    income
-   4: Concept not applicable (territories or reserves) - in low income

According to the census code book, the variable 'LOLICOA' has the
following possible values:[\^]

[\^] Refers to the income situation of a person in relation to the
Low-income cut-offs, after-tax (LICO-AT). Persons with income that is
below this low-income line are considered to be in low income.

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Concept not applicable (territories or reserves) - not in low
    income
-   4: Concept not applicable (territories or reserves) - in low income

According to the census code book, the variable 'LoLIMA' has the
following possible values:[\^]

[\^] Refers to the income situation of a person in relation to the
Low-income measure, after-tax (LIM-AT). Persons with income that is
below this low-income line are considered to be in low income.

-   -3: Concept not applicable (collective households)
-   1: Not in low income
-   2: In low income
-   3: Not in low income (territories or reserves)
-   4: In low income (territories or reserves)
-   5: Concept not applicable (collective households) - not in low
    income
-   6: Concept not applicable (collective households) - in low income

We will use the variable *ACTLMFLG - Activities of daily living: Flag*
as an indicative of disability, because the Census of Population does
not have a variable regarding disability. The question on activities of
daily living refers to difficulties a person may have doing certain
activities as a result of physical, cognitive, mental, or other
health-related conditions or problems. We will select all those
respondents who affirm to have frequent (always) difficulties to develop
regular activities. According to the census code book, the variable
'ACTLMFLG' has the following possible values:

-   -2: Not stated
-   1: Yes, always
-   2: Yes, often
-   3: Yes, sometimes
-   4: No

According to the census code book, the variable 'CFAMSTSIMPLE' has the
following possible values:

-   1: Married spouse or common-law partner
-   2: Parent in a one-parent family
-   3: Child
-   4: Person not in a census family

According to the census code book, the variable 'IMMCAT5' has the
following possible values:

-   1: Non-immigrants
-   2: Non-permanent residents3 Immigrants admitted before 1980
-   100010: Economic immigrants - Principal applicants
-   100020: Economic immigrants - Secondary applicants
-   200000: Immigrants sponsored by family
-   300000: Refugees
-   400000: Other immigrants

According to the census code book, "Visible minority" refers to whether
a person is a visible minority as defined by the Employment Equity Act.
The Employment Equity Act defines visible minorities as "persons, other
than Aboriginal peoples, who are non-Caucasian in race or non-white in
colour." The visible minority population consists mainly of the
following groups: South Asian, Chinese, Black, Filipino, Arab, Latin
American, Southeast Asian, West Asian, Korean and Japanese. The variable
'DVisMin' has the following possible values:

-   1: Other (not a visible minority)
-   2: South Asian
-   3: Chinese
-   4: Black
-   5: Filipino
-   6: Arab
-   7: Latin American
-   8: Southeast Asian
-   9: West Asian
-   10: Korean
-   11: Japanese
-   12: Visible minority, n.i.e.
-   13: Multiple visible minorities
-   14: Indigenous peoples

In the next chunk, we will rename the travel modes to facilitate the
readability of the data.

```{r labelling-variables}
census_filtered_labeled <- census_filtered %>%
                   mutate( # Labour force
                     LBR_FORC_label = 
                            case_when(LBR_FORC == 1 ~ "Employed",
                                      LBR_FORC == 2 ~ "Unemployed"),
                          LBR_FORC_label = factor(LBR_FORC_label), 
                     # FPTim: Labour: Full-time or part-time weeks worked in 2020
                     FPTim_label = 
                            case_when(FPTim == -5 ~ "Unemployed",
                                      FPTim == 1 ~ "Full-time worker",
                                      FPTim == 2 ~ "Part-time worker"),
                          FPTim_label = factor(FPTim_label, levels = c("Full-time worker", "Part-time worker", "Unemployed")), 
                     # Gender 
                          GENDER_BINARY_label = 
                            case_when(GENDER_BINARY == 1 ~ "Woman+",
                                      GENDER_BINARY == 2 ~ "Man+"),
                          GENDER_BINARY_label = factor(GENDER_BINARY_label, levels = c("Man+", "Woman+")),
                     # Low income population - LOLICOA
                     LOLICOA_label = 
                       case_when(LOLICOA == -3 ~ "Not in low income", 
                                 LOLICOA == 1 ~ "Not in low income",
                                 LOLICOA == 2 ~ "In low income",
                                 LOLICOA == 3 ~ "Not in low income",
                                 LOLICOA == 4 ~ "In low income"),
                     LOLICOA_label = factor(LOLICOA_label), 
                    LoLIMA_label = 
                       case_when(LoLIMA == -3 ~ "Not in low income", 
                                 LoLIMA == 1 ~ "Not in low income",
                                 LoLIMA == 2 ~ "In low income",
                                 LoLIMA == 3 ~ "Not in low income",
                                 LoLIMA == 4 ~ "In low income",
                                 LoLIMA == 5 ~ "Not in low income",
                                 LoLIMA == 6 ~ "In low income"),
                     LoLIMA_label = factor(LoLIMA_label), 
                    Low_Income_label = 
                      case_when(LoLIMA_label == "In low income" | LOLICOA_label == "In low income" ~ "In low income",
                                TRUE ~ "Not in low income"), 
                    Low_Income_label = factor(Low_Income_label), 
                    # Disability
                    ACTLMFLG_label = 
                       case_when(ACTLMFLG == -2 ~ "Without frequent difficulties", 
                                 ACTLMFLG == 1 ~ "With frequent difficulties",
                                 ACTLMFLG == 2 ~ "Without frequent difficulties",
                                 ACTLMFLG == 3 ~ "Without frequent difficulties",
                                 ACTLMFLG == 4 ~ "Without frequent difficulties"),
                    ACTLMFLG_label = factor(ACTLMFLG_label, levels = c("Without frequent difficulties", "With frequent difficulties")), 
                    # Single-parent family
                    CFAMSTSIMPLE_label = 
                       case_when(CFAMSTSIMPLE == 1 ~ "Non single-parent family",
                                 CFAMSTSIMPLE == 2 ~ "Single-parent family",
                                 CFAMSTSIMPLE == 3 ~ "Non single-parent family",
                                 CFAMSTSIMPLE == 4 ~ "Non single-parent family"),
                     CFAMSTSIMPLE_label = factor(CFAMSTSIMPLE_label, levels = c("Non single-parent family", "Single-parent family")), 
                    # Refugees
                    IMMCAT5_label = 
                       case_when(IMMCAT5 == 300000 ~ "Refugee",
                                 IMMCAT5 != 300000 ~ "Not a refugee"),
                     IMMCAT5_label = factor(IMMCAT5_label), 
                    # New immigrant
                    PerImmA_label = 
                       case_when(PerImmA == 10 ~ "New immigrant",
                                 PerImmA != 10 ~ "Not a new immigrant"),
                     PerImmA_label = factor(PerImmA_label),
                    # Visible Minority (binary)
                    DVisMin_binary_label = 
                       case_when(DVisMin == 1 ~ "Not a visible minority",
                                 DVisMin != 1 ~ "Visible minority"),
                     DVisMin_binary_label = factor(DVisMin_binary_label, 
                                                   levels = c("Not a visible minority","Visible minority")), 
                    # Visible Minority (full-group)
                    DVisMin_label = 
                      case_when(DVisMin == 1  ~ "Other (not a visible minority)",
                                DVisMin == 2  ~ "South Asian",
                                DVisMin == 3  ~ "Chinese",
                                DVisMin == 4  ~ "Black",
                                DVisMin == 5  ~ "Filipino",
                                DVisMin == 6  ~ "Arab",
                                DVisMin == 7  ~ "Latin American",
                                DVisMin == 8  ~ "Southeast Asian",
                                DVisMin == 9  ~ "West Asian",
                                DVisMin == 10 ~ "Korean",
                                DVisMin == 11 ~ "Japanese",
                                DVisMin == 12 ~ "Visible minority, n.i.e.",
                                DVisMin == 13 ~ "Multiple visible minorities",
                                DVisMin == 14 ~ "Indigenous peoples"),
                    DVisMin_label = factor(DVisMin_label),
                    # AGE
                    AGE_label = 
                      case_when(AGE <= 24 ~ "Youth",
                                TRUE  ~ "Adult"),
                    AGE_label = factor(AGE_label), 
                    # Number of weeks worked in 2020 
                    WEEKS = 
                      case_when(WEEKS < 0 ~ 0, 
                                TRUE ~ WEEKS), 
                    # Low educated 
                    HCDD_label =
                      case_when(HCDD == 1 ~ "Low educated", 
                                TRUE ~ "High-school diploma"),
                    HCDD_label = factor(HCDD_label, levels = c("High-school diploma", 
                                                               "Low educated")),
                    # Looked for paid work in the previous month 
                    LOKWK = 
                      case_when(LOKWK == -5 ~ "Employed",
                                LOKWK == 1 ~  "No",
                                LOKWK == 2 ~  "Yes (for full-time work)",
                                LOKWK == 3 ~ "Yes (for part-time work)"),
                    LOKWK = factor(LOKWK, levels = c("Yes (for full-time work)",
                                                     "Yes (for part-time work)", 
                                                     "No", 
                                                     "Employed")))
```

View summary statistics from the data frame:

```{r census-summary}
summary(census_filtered_labeled)
```

## Reading the job accessibility and spatial filter indicators 

Setting the folder with the accessibility measures and spatial filter: 

```{r read-accessibility}
# Folder  
measures_folder <- paste0(here::here(),"/data-raw/output/PCD3520/DA/spatial-filter-measures/")

# Read file
SA_mode <- read.csv(paste0(measures_folder, "spatial_filter_original.csv"))
```

```{r}
SA_mode_wide <- SA_mode %>% 
  rename(SA = SA_im) %>%
  pivot_wider(id_cols = PRCDDA,
              names_from = PwMode,
              values_from = c(SA, Sf),
              names_sep = "_")
```


## Reading the tables with the disadvantaged population groups 

Setting the folder with the disadvantaged population groups: 

```{r read-disadvantaged-files}
# Folder  
measures_folder <- paste0(here::here(),"/data-raw/output/PCD3520/DA/disadvantaged-tables/")

# Read file
dpg_pvalues <- read.csv(paste0(measures_folder, "pvalue_table_release.csv"))
dpg_table <- read.csv(paste0(measures_folder, "summary_pop_table_release.csv"))
dvm_table <- read.csv(paste0(measures_folder, "summary_vmin_table_release.csv"))
```

Defining if the respondent is part a disadvantaged group: 

```{r creating-equity-variable}
significant_dpg <- dpg_pvalues %>%
  filter(Significance == 'Significant', 
         Variable != "DVisMin_binary_label")

real_dvm <- dvm_table %>% 
  filter(Equity == "Disadvantaged", 
         Median_rate < 1)

logic_matrix <- map2(significant_dpg$Variable,
                     significant_dpg$Disadvantaged,
                     ~ census_filtered_labeled[[.x]] == .y)

logic_df <- as.data.frame(logic_matrix)

census_filtered_labeled <- census_filtered_labeled %>%
  mutate(Equity = if_else(
    apply(logic_df, 1, any),
    "Disadvantaged",
    "Non-disadvantaged"
  ))

# For visible minorities, we will select only those groups that in fact are disadvantaged (not necessarily all groups within the visible minority group is considered as disadvantaged)

if ("DVisMin_binary_label" %in% dpg_pvalues$Variable) {

    logic_matrix <- map2(
    real_dvm$Variable,
    real_dvm$Category,
    ~ census_filtered_labeled[[.x]] == .y
  )

  logic_df <- as.data.frame(logic_matrix)

  census_filtered_labeled <- census_filtered_labeled %>%
    mutate(Equity = if_else(
      apply(logic_df, 1, any),
      "Disadvantaged",
      "Non-disadvantaged"
    ))
}
```

## Creating the tables to build the regression models

```{r respondents}
respondents <- census_filtered_labeled %>%
  left_join(SA_mode_wide, by = "PRCDDA")
```

```{r dis-nondis}
disadvantaged <- respondents %>%
  filter(Equity == "Disadvantaged")

non_disadvantaged <- respondents %>%
  filter(Equity == "Non-disadvantaged")
```  
